---
title: "snRNA-seq TSC2: Case-control analysis - Neurons - GWAS Workflow"
output: html_notebook
author: "Frederik Nørby Friis Sørensen, University of Copenhagen"
date: "April 2026"
---

##LIBS
```{r setup}
library(dplyr)
library(magrittr)
library(conos)
library(pagoda2)
library(scHelper)
library(cacoa)
library(ggplot2)
library(ggrastr)
library(cowplot)
library(HDF5Array)
library(Seurat)
set.seed(1)
```

# Cellect/Cellex GWAS and snRNA-seq integration workflow

## Load data
```{r}
annotation_n <- readRDS("../../../TSC2024_data_files/5.neurons/annotation_neurons.rds")
con_neurons <- readRDS("../../5.neurons/con_neurons.rds")
cm.merged.raw <- con_neurons$getJointCountMatrix(raw=T)
```


## Convert genenames to ENSEMBL
```{r}
library(org.Hs.eg.db)
```

```{r}
gnames <- clusterProfiler::bitr(geneID = colnames(cm.merged.raw), fromType = "SYMBOL", toType = "ENSEMBL", OrgDb = org.Hs.eg.db)
gnames %<>% .[match(unique(.$SYMBOL), .$SYMBOL),] %>% .[match(unique(.$ENSEMBL), .$ENSEMBL),]

cm.merged.raw %<>% .[,match(gnames$SYMBOL, colnames(.))]
colnames(cm.merged.raw) <- gnames$ENSEMBL
```

## Make annotation table

```{r}
cell.groups <- annotation_n$l4 %>%
  as.character() %>%
  setNames(names(annotation_n$l4)) %>%
  .[match(rownames(cm.merged.raw),names(.))] %>%
  cbind()%>% 
  data.frame(stringsAsFactors = F) %>% 
  tibble::rownames_to_column() %>% 
  setNames(c("cell_id","cell_type"))
cell.groups$cell_id %<>% gsub(".", "", ., fixed = T)
cell.groups$cell_id %<>% gsub("!", "", ., fixed = T)
```

## Extract TSC2 cells

```{r}
sample.per.cell <- con_neurons$getDatasetPerCell()
sample.groups = c(rep("Control",6), rep("TSC2",11)) %>% setNames(names(con_neurons$samples))
ep.cells <- sample.per.cell[sample.per.cell %in% names(sample.groups[sample.groups == "TSC2"])] %>% factor()
names(ep.cells) <- names(ep.cells) %>% gsub("!","", ., fixed = T)
ep.cells %<>% factor()
groups.ep <- cell.groups %>% .[.$cell_id %in% names(ep.cells),]
rownames(cm.merged.raw) %<>% gsub("!", "", ., fixed = T)
ep.cm <- cm.merged.raw %>% .[rownames(.) %in% groups.ep$cell_id,]
dim(ep.cm); dim(groups.ep) #check if dimensions of matrix and cells are fine (only disease cells are included)
```

## Write files for cellex processing
```{r}
library(HDF5Array)
library(Matrix)

dir.create("CELLEX")
writeHDF5Array(ep.cm, "CELLEX/ep.h5", "data", verbose = T)
write.csv(ep.cm %>% colnames(), "CELLEX/genes.ep.csv", row.names = F)
write.csv(ep.cm %>% rownames(), "CELLEX/cells.ep.csv", row.names = F)
write.csv(groups.ep, "CELLEX/metadata.ep.csv", row.names = F)
```

## Run the python script below in conda env with cellex and prerequisites installed.

```{python}
import numpy as np
import pandas as pd
import cellex
# this script was developed using pybedtools v. 0.8.0import h5py

data = pd.DataFrame(np.array(h5py.File("ep.h5")['data']), dtype="float32")
data.columns = pd.read_csv("cells.ep.csv").values
data.columns = [v[0] for v in data.columns]
data.index = pd.read_csv("genes.ep.csv").values
data.index = [v[0] for v in data.index]

metadata = pd.read_csv("metadata.ep.csv", index_col=0)

eso = cellex.ESObject(data=data, annotation=metadata, verbose=True)
eso.compute(verbose=True)

eso.results["esmu"].to_csv("Cellex.ep.results.csv.gz")
```

## When cellex results are ready and GWAS summary stats are munged, proceed to CELLECT workflow to obtain prioritization results
## If you decide not to run the pipeline yourself, prioritization results are in the folder under MAGMA_results

##Focal Epilepsy
###MAGMA

```{r}
cell_types <- c(
  "L2_CUX2_LAMP5_PDGFD","L2_3_CUX2_FREM3_NTNG1", "L3_CUX2_SPOCK3_PRSS12", 
  "L4_RORB_SCHLAP1_ARHGAP15", "L4_RORB_SCHLAP1_MET", "L4_RORB_SCHLAP1_MME", 
  "L5_6_FEZF2_LRRK1_PCP4", "L5_6_FEZF2_LRRK1_SEMA3E", "L5_6_FEZF2_TLE4_ABO", 
  "L5_6_FEZF2_TLE4_HTR2C", "L5_6_FEZF2_TLE4_SCUBE1", "L5_6_THEMIS_NR4A2_NTNG2", 
  "L5_6_THEMIS_SEMA3A", "ID2_CHRNA7_CNR1", "ID2_CHRNA7_NCKAP5", "ID2_CHRNA7_PAX6",
  "ID2_LAMP5_CRH", "ID2_LAMP5_MEIS2", "ID2_LAMP5_NMBR", "ID2_LAMP5_NOS1", 
  "VIP_ABI3BP", "VIP_CBLN1", "VIP_CRH", "VIP_NRG1", "VIP_SEMA3C", "VIP_TRPM3_SSTR1", "VIP_TYR",
  "PVALB_BTBD11", "PVALB_LGR5", "PVALB_MEPE", "PVALB_UNC5B",
  "SST_CALB1", "SST_ISOC_THSD7B", "SST_NOS1", "SST_NPY", "SST_STK32A", "SST_TAC3", "SST_TH"
)

dat.ld <- read.csv("MAGMA_results/prioritization_focal.csv")
dat.ld$neg_logP <- -log10(dat.ld$pvalue)

dat.ld <- dat.ld %>%
  dplyr::mutate(type = "LD") %>%
  dplyr::filter(annotation %in% cell_types) %>%
  dplyr::mutate(annotation = factor(annotation, levels = cell_types))

p.cutoff <- 0.05 / nrow(dat.ld)

cellect <- ggplot(dat.ld, aes(annotation, pvalue, col = annotation)) + 
  geom_point() +
  labs(x = "", y = "P value") +
  geom_hline(yintercept = 0.05, col = "black", linetype = "dotted") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90),
    legend.position = "none",
    axis.text.y = element_text(colour = ifelse(dat.ld$pvalue <= p.cutoff, "red", "black"))
  ) +
  coord_flip() + 
  scale_color_manual(values = dat.ld$pvalue %>% sapply(function(x) {
    if (x <= p.cutoff) return("red")
    else if (x > p.cutoff & x <= 0.05) return("gray")
    else return("black")
  }) %>% setNames(as.character(dat.ld$annotation)))
```


```{r}
cellect

filename = "../figures/Cellect_GWAS_Focal_epilepsy_MAGMA.pdf"
pdf.width = 5
pdf.height = 6
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```
# Fig 6e
```{r}
cellect_log <- ggplot(dat.ld, aes(annotation, neg_logP, col = annotation)) + 
  geom_point() +
  labs(x = "", y = "-log10(p)") +
  geom_hline(yintercept = 1.301, col = "black", linetype = "dotted") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90),
    legend.position = "none"
  ) +
  coord_flip() + 
  scale_color_manual(values = dat.ld$neg_logP %>% sapply(function(x) {
    if (x > -log10(p.cutoff)) return("red")
    else if (x < -log10(p.cutoff) & x > 1.301) return("gray")
    else return("black")
  }) %>% setNames(as.character(dat.ld$annotation)))
```

```{r fig.height=6.5, fig.width=6, message=FALSE}
cellect_log

filename = "../figures/Cellect_GWAS_Focal_epilepsy_Logscale_MAGMA.pdf"
pdf.width = 6
pdf.height = 6.5
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)

```

## Generalized
###MAGMA
```{r}
  dat.ld <- read.csv("MAGMA_results/prioritization_generalized.csv")
dat.ld$neg_logP <- -log10(dat.ld$pvalue)

dat.ld <- dat.ld %>%
  dplyr::mutate(type = "LD") %>%
  dplyr::filter(annotation %in% cell_types) %>%
  dplyr::mutate(annotation = factor(annotation, levels = cell_types))

p.cutoff <- 0.05 / nrow(dat.ld)

cellect <- ggplot(dat.ld, aes(annotation, pvalue, col = annotation)) + 
  geom_point() +
  labs(x = "", y = "P value") +
  geom_hline(yintercept = 0.05, col = "black", linetype = "dotted") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90),
    legend.position = "none",
    axis.text.y = element_text(colour = ifelse(dat.ld$pvalue <= p.cutoff, "red", "black"))
  ) +
  coord_flip() + 
  scale_color_manual(values = dat.ld$pvalue %>% sapply(function(x) {
    if (x <= p.cutoff) return("red")
    else if (x > p.cutoff & x <= 0.05) return("gray")
    else return("black")
  }) %>% setNames(as.character(dat.ld$annotation)))
```

```{r}
cellect

filename = "../figures/Cellect_GWAS_generalized_epilepsy_MAGMA.pdf"
pdf.width = 5
pdf.height = 6
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

```{r}
cellect_log <- ggplot(dat.ld, aes(annotation, neg_logP, col = annotation)) + 
  geom_point() +
  labs(x = "", y = "-log10(p)") +
  geom_hline(yintercept = 1.301, col = "black", linetype = "dotted") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90),
    legend.position = "none"
  ) +
  coord_flip() + 
  scale_color_manual(values = dat.ld$neg_logP %>% sapply(function(x) {
    if (x > -log10(p.cutoff)) return("red")
    else if (x < -log10(p.cutoff) & x > 1.301) return("gray")
    else return("black")
  }) %>% setNames(as.character(dat.ld$annotation)))
```
# Fig 6f
```{r fig.height=6.5, fig.width=6, message=FALSE}
cellect_log
filename = "../figures/Cellect_GWAS_generalized_epilepsy_Logscale_MAGMA.pdf"
pdf.width = 6
pdf.height = 6.5
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)

```

```{r}
sessionInfo()
```

