---
title: "snRNA-seq TSC2 preprocessing with Cellbender"
output: html_notebook
author: "Frederik Nørby Friis Sørensen, University of Copenhagen"
date: "April 2026"
---

### Setup libraries
```{r, setup}
library(dplyr)
library(magrittr)
library(conos)
library(pagoda2)
library(cacoa)
library(reticulate)
library(ggplot2)
library(ggrastr)
library(cowplot)
library(HDF5Array)
library(Seurat)
library(CRMetrics)
library(ggrastr)
set.seed(1)
```

```{r message=FALSE, include=FALSE}
# Check if 'data' folder exists, if not, create it
if (!dir.exists("data")) {
  dir.create("data")
  cat("Created 'data' folder\n")
} else {
  cat("'data' folder already exists\n")
}

# Check if 'figure' folder exists, if not, create it
if (!dir.exists("figures")) {
  dir.create("figures")
  cat("Created 'figures' folder\n")
} else {
  cat("'figures' folder already exists\n")
}

```

# CRMetrics
## Make object and load count matrices
### Replace data path with the location you store the count matrices. 
```{r}
data_path= "../../../Sequencing/counts"
crm <- CRMetrics$new(data.path = data_path, n.cores = 100)
```

```{r}
crm$selectMetrics()
```
```{r}
mean_reads <- crm$plotSummaryMetrics(comp.group = "sample", metrics = "mean reads per cell", plot.geom = "bar")
mean_reads$data
mean(mean_reads$data$value)
total_cells <- crm$plotSummaryMetrics(comp.group = "sample", metrics = "estimated number of cells", plot.geom = "bar")
total_cells$data
sum(total_cells$data$value)
```
```{r fig.height=12, fig.width=12}
metrics.to.plot <- crm$selectMetrics(ids = c(1:4,6,18,19,11,15))
p <- crm$plotSummaryMetrics(metrics = metrics.to.plot, 
                       plot.geom = "bar")+theme_minimal()
p
ggsave(plot = p, filename = "figures/all_cells/cellranger_metrics_precb.pdf",dpi = 300, width = 12,height = 12,units = "in")
```
```{r}
crm$selectMetrics()
```

```{r fig.height=9, fig.width=9}
crm$addComparison("sample")
```


## Here cellbender scripts will be generated for use in terminal to run the pipeline
### Pipeline is run in conda environment in terminal.
```{r fig.height=15, fig.width=20}
#crm$prepareCellbender(n.cores = 100)
```
```{r}
#crm$saveCellbenderScript()
```

## Before proceeding to this step below, Cellbender unsupervised ambient RNA removal needs to be performed.

```{r}
crm$cms <- NULL
crm$detailed.metrics <- NULL
crm$cms.preprocessed <- NULL
crm$con <- NULL
# Set data path to your cellbender counts output folder (if you run it yourself, if not use the data file provided)
crm$addCms(cellbender = T,data.path = "../../TSC2024_data_files/0.cellbender_counts/")
crm$addDetailedMetrics(n.cores = 100)
```


```{r}
mean_reads <- crm$plotSummaryMetrics(comp.group = "sample", metrics = "mean reads per cell", plot.geom = "bar")
mean_reads$data
mean(mean_reads$data$value)
mean_reads
total_cells <- crm$plotSummaryMetrics(comp.group = "sample", metrics = "estimated number of cells", plot.geom = "bar")
total_cells$data
sum(total_cells$data$value)
total_cells
```

```{r fig.height=4, fig.width=8}
metrics.to.plot <- crm$detailed.metrics$metric %>%
  unique()

umi <- crm$plotDetailedMetrics(metrics = metrics.to.plot[1], 
                        plot.geom = "violin")
gene_count <- crm$plotDetailedMetrics(metrics = metrics.to.plot[2], 
                        plot.geom = "violin")
median(umi$data$value)
median(gene_count$data$value)
```


```{r fig.height=12, fig.width=12}
library(ggplot2)
library(ggforce)
p <- crm$plotCbCells(data.path ="../../TSC2024_data_files/0.cellbender_counts/") +
  theme(axis.text.x = element_text(angle = 45,   hjust = 1, vjust = 1),
                       strip.background = element_blank(), # Remove the grey header box
                       strip.text = element_text(face = "plain"))
p
ggsave(plot = p, filename = "figures/all_cells/cellbender_metrics.pdf",dpi = 300, width = 12,height = 12,units = "in")
```


```{r fig.height=10, fig.width=10}
p <- crm$plotCbCellProbs(data.path ="../../TSC2024_data_files/0.cellbender_counts/") %>% rasterize(dpi=600) +
  theme(axis.text.x = element_text(angle = 45,   hjust = 1, vjust = 1),
                       strip.background = element_blank(), # Remove the grey header box
                       strip.text = element_text(face = "plain"))

p
ggsave(plot = p, filename = "figures/all_cells/cellbender_cell_probabilities.pdf",dpi = 300, width = 12,height = 12,units = "in")
```


```{r fig.height=8, fig.width=12}
p <- crm$plotCbAmbExp(data.path ="../../TSC2024_data_files/0.cellbender_counts/") %>% rasterize(dpi=600)+
  theme(axis.text.x = element_text(angle = 45,   hjust = 1, vjust = 1),
                       strip.background = element_blank(), # Remove the grey header box
                       strip.text = element_text(face = "plain"))
p
ggsave(plot = p, filename = "figures/all_cells/cellbender_amb_expression.pdf",dpi = 300, width = 12,height = 12,units = "in")
```

```{r fig.height=6, fig.width=6}
p <- crm$plotCbAmbGenes(data.path ="../../TSC2024_data_files/0.cellbender_counts/")+
  theme(axis.text.x = element_text(angle = 45,   hjust = 1, vjust = 1),
                       strip.background = element_blank(), # Remove the grey header box
                       strip.text = element_text(face = "plain"))
p
ggsave(plot = p, filename = "figures/all_cells/cellbender_amb_genes.pdf",dpi = 300, width = 6,height = 6,units = "in")
```

```{r fig.height=16, fig.width=16, paged.print=TRUE}
p <- crm$plotCbTraining(data.path ="../../TSC2024_data_files/0.cellbender_counts/")+
  theme(axis.text.x = element_text(angle = 45,   hjust = 1, vjust = 1),
                       strip.background = element_blank(), # Remove the grey header box
                       strip.text = element_text(face = "plain"))
p
ggsave(plot = p, filename = "figures/all_cells/cellbender_training_curves.pdf",dpi = 300, width = 12,height = 12,units = "in")
```

## Proceed to normal filtering and preprocessing:
```{r include=FALSE}
crm$doPreprocessing(n.cores = 100)
```
```{r include=FALSE}
crm$createEmbedding(n.cores = 100)
```

### General embedding
```{r fig.height=8, fig.width=8}
crm$con$plotGraph(size=0.1, shuffle.colors=T, alpha=0.3, font.size=3) %>% rasterise(dpi=600) +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank(), panel.border = element_blank())
filename = "figures/all_cells/UMAP_all_cells_no_filtering.pdf"
pdf.width = 10
pdf.height = 10
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

## Depth filtering 

```{r}
depth <- getConosDepth(crm$con)
ggplot(data.frame(depth) %>% setNames(c("x")), aes(x)) + geom_histogram(binwidth = 0.005) + xlim(c(0,1e6)) + scale_x_log10() + geom_vline(xintercept = 1.7e3, color="red") + labs(title="Depth Profile") + xlab("Cell size (UMIs)") + ylab("Number of cells")
filename = "figures/all_cells/depth_profile_all_cells_pre_filter.pdf"
pdf.width = 10
pdf.height = 8
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

### Sample per cell 
```{r fig.height=8, fig.width=10}
dspc <- crm$con$getDatasetPerCell()
```
```{r fig.height=8, fig.width=10}
crm$con$plotGraph(groups=dspc, show.legend=T, size=0.1, alpha=0.3,font.size=0) %>% rasterise(dpi=600)  +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank(), panel.border = element_blank())
filename = "figures/all_cells/dspc_all_cells_pre_filter.pdf"
pdf.width = 11.5
pdf.height = 10
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```


## Doublet detection 
```{r eval=FALSE, include=FALSE}
library(reticulate)
conda_create("r-reticulate", 
             conda = "/opt/software/miniconda/4.12.0/condabin/conda", 
             python_version = 3.8)
conda_install("r-reticulate", 
              conda = "/opt/software/miniconda/4.12.0/condabin/conda", 
              pip = TRUE, 
              packages = c("scrublet","doubletdetection"))
```
### Scrublet
```{r}
crm$detectDoublets(conda.path = "/opt/software/miniconda/4.12.0/condabin/conda",method = "scrublet",env = "r-reticulate", verbose = T)
#Detected 3444 possible doublets out of 122731 cells.
```

```{r fig.height=8, fig.width=8}
crm$plotEmbedding(doublet.method = "scrublet") %>% rasterise(dpi=600) +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank(), panel.border = element_blank())
filename = "figures/all_cells/scrublet_labels_all_cells.pdf"
pdf.width = 10
pdf.height = 10.2
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)

```
```{r fig.height=8, fig.width=8}
crm$plotEmbedding(doublet.method = "scrublet", 
             doublet.scores = TRUE) %>% rasterise(dpi=600)  +
    theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank(), panel.border = element_blank())
filename = "figures/all_cells/scrublet_scores_all_cells.pdf"
pdf.width = 10
pdf.height = 10.2
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```
### DoubletDetection (not run for this dataset)
```{r eval=FALSE, include=FALSE}
crm$detectDoublets(conda.path = "/opt/software/miniconda/4.12.0/condabin/conda",env = "r-reticulate",method = "DoubletDetection", verbose = T)
```
```{r eval=FALSE, include=FALSE}
crm$plotEmbedding(doublet.method = "doubletdetection") +
    theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank(), panel.border = element_blank())
filename = "figures/all_cells/doubletdetection_labels_all_cells.pdf"
pdf.width = 10
pdf.height = 10.2
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)

```

## Mitochondrial gene fraction 
```{r fig.height=8, fig.width=8}
crm$plotEmbedding(mito.frac = TRUE, 
             mito.cutoff = 0.05, 
             species = "human") %>% rasterise(dpi=600) +
    theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank(), panel.border = element_blank())
filename = "figures/all_cells/mitofraction_labels_all_cells.pdf"
pdf.width = 10
pdf.height = 10.2
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```
```{r fig.height=6, fig.width=10}
group.per.cell <- dspc %>% as.character()
mito_frac <- mitoFraction(crm$con, "human")
ggplot(data.frame(Mito=mito_frac, Group=group.per.cell), aes(Group, Mito)) + ggbeeswarm::geom_quasirandom(size=0.1) %>% rasterise(dpi=600) +
  geom_hline(yintercept = 0.05, col="red")+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  ylab("Mitochondrial gene fraction") +
  xlab("Sample") +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank(), panel.border = element_blank())
filename = "figures/all_cells/mitofraction_sample_scatter_all_cells.pdf"
pdf.width = 10
pdf.height = 8
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```


## Scrublet filter plots
```{r fig.height=8, fig.width=10}
crm$plotFilteredCells(type="embedding",
                      doublet.method = "scrublet", 
                      depth = T,
                      depth.cutoff = 100,
                      mito.frac = TRUE, 
                      mito.cutoff = 0.05, 
                      species = "human",
                      cols = c("aliceblue","deeppink3","darkblue","orange","tan2","darkcyan","black","chartreuse1"),
                      size = 0.4) %>% rasterise(dpi=600)  +
    theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank(), panel.border = element_blank())
filename = "figures/all_cells/cellstofilter_all_cells_prefilter_embedding.pdf"
pdf.width = 11.5
pdf.height = 10.2
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```
```{r fig.height=8, fig.width=10}
crm$plotFilteredCells(type = "bar", 
                      doublet.method = "scrublet", 
                      depth = FALSE,
                      depth.cutoff = 0,
                      mito.frac = TRUE, 
                      mito.cutoff = 0.05, 
                      species = "human") +
    theme(panel.border = element_blank())
filename = "figures/all_cells/cellstofilter_all_cells_prefilter_barplot.pdf"
pdf.width = 11.5
pdf.height = 10
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

```{r}
saveRDS(crm, "data/crm_ready_filtering_manuscript.rds", compress=F)
```
## First filtering
### Here only doublets and cells with 5% > mitochondrial gene fraction are filtered. 

```{r}
crm$filterCms(mito.cutoff =0.05,
              species ="human",
              doublets="scrublet")
```

```{r}
saveRDS(crm$cms.filtered, "data/cms_doublet_mito_filtered_all_cells.rds", compress=F)
```

## Make Conos object
```{r eval=FALSE, include=FALSE}
con<- quickConos(crm$cms.filtered,
                  names(crm$cms.filtered), 
                  make.geneknn = F,
                  get.largevis = F,
                  n.odgenes = 3e3,
                  nPcs = 100,
                  min.cells.per.gene = 3, 
                  min.transcripts.per.cell = 200, 
                  ncomps = 40,
                  get.tsne = F,
                  n.cores.p2=100,
                  n.cores.con=100)
con <- con$con

saveRDS(object = con, file = "data/con_doublets_mito_filterd_all_cells_manuscript.rds", compress = F)
```
```{r eval=FALSE, include=FALSE}
saveRDS(crm, "data/crm_doublet_mito_filtered_all_cells_manuscript.rds", compress=F)
```

```{r}
crm <- readRDS("data/crm_doublet_mito_filtered_all_cells_manuscript.rds")
con <- readRDS("data/con_doublets_mito_filterd_all_cells_manuscript.rds")
anno <- readRDS("anno_for_filtering.rds")
```


```{r fig.height=8, fig.width=8}
con$plotGraph(size=0.1,
              alpha=0.1,
              groups = anno,
              font.size=6,
              shuffle.colors=T) %>% rasterise(dpi=600)  +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(),
          panel.border = element_blank())


filename = "figures/all_cells/UMAP_mito_doublet_filter_anno.pdf"
pdf.width = 10
pdf.height = 10
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```


```{r fig.height=12, fig.width=12}
c("RBFOX3","SLC17A7","GAD1","AQP4","VCAN","MOG","CSF1R","PTPRC","CLDN5") %>%
lapply(function(g) con$plotGraph(gene=g,
                                 title=g,
                                 size=0.2,
                                 alpha=1) %>% rasterise(dpi=600) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(),
          panel.border = element_blank())) %>%
  cowplot::plot_grid(plotlist=.,ncol=3)  

filename = "figures/all_cells/UMAP_marker_genes.pdf"
pdf.width = 15
pdf.height = 15
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)

```

# Split neurons and glia for further preprocessing

## Neuron subset
```{r}
cms_neurons <- crm$cms.filtered
cms_neurons %<>% lapply(function(x) x[,colnames(x) %in% names(anno[anno %in% c("neurons")])])
```
```{r}
# Define the function to modify column names
modify_colnames <- function(matrix) {
  # Modify the column names by removing parts of the string using gsub
  colnames(matrix) <- gsub("^.*!!", "", colnames(matrix))
  return(matrix)
}

# Apply this function to each matrix in the list using sapply
cms_neurons_clean <- sapply(cms_neurons,
                            modify_colnames,
                            simplify = FALSE)
```

## Glia subset
```{r}
cms_glia <- crm$cms.filtered
cms_glia %<>% lapply(function(x) x[,colnames(x) %in% names(anno[anno %in% c("glia")])])
```

```{r}
# Define the function to modify column names
modify_colnames <- function(matrix) {
  # Modify the column names by removing parts of the string using gsub
  colnames(matrix) <- gsub("^.*!!", "", colnames(matrix))
  return(matrix)
}

# Apply this function to each matrix in the list using sapply
cms_glia_clean <- sapply(cms_glia,
                         modify_colnames,
                         simplify = FALSE)
```

# Glia


```{r}
crm_glia<- CRMetrics$new(cms = cms_glia_clean,
                         n.cores = 100) #Remember to =F unique names as it will introduce problems down the line with plots if cell names are made unique twice. 
```

```{r}
crm_glia$addDetailedMetrics(n.cores = 100)
```
```{r}
dir.create("figures/glia")
```

```{r fig.height=4, fig.width=8}
metrics.to.plot <- crm_glia$detailed.metrics$metric %>%
  unique()

crm_glia$plotDetailedMetrics(metrics = metrics.to.plot, 
                        plot.geom = "violin")+
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(),
          panel.border = element_blank())

filename = "figures/glia/violin_plot_metrics.pdf"
pdf.width = 10
pdf.height = 5
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```
```{r}
crm_glia$doPreprocessing(n.cores = 100)
```
```{r}
crm_glia$createEmbedding(n.cores = 100)
```

```{r}
saveRDS(crm_glia, "data/crm_glia_cb_prefilter_manuscript.rds", compress=F)

```

      #General embedding
```{r fig.height=8, fig.width=8}
crm_glia$con$plotGraph(size=0.1,
                       shuffle.colors=T,
                       alpha=0.5) %>% rasterise(dpi=600) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(),
          panel.border = element_blank())

filename = "figures/glia/UMAP_glia.pdf"
pdf.width = 10
pdf.height = 10
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)

```
```{r}
c("AQP4","VCAN","MOG","CSF1R","PTPRC","CLDN5") %>%
lapply(function(g) crm_glia$con$plotGraph(gene=g,
                                          title=g,
                                          size=0.2,
                                          alpha=1) %>% rasterise(dpi=300) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(),
          panel.border = element_blank())) %>%
  cowplot::plot_grid(plotlist=.,ncol=3)

filename = "figures/glia/UMAP_glia_marker_genes.pdf"
pdf.width = 10
pdf.height = 6.66
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

## Depth filtering 

```{r}
crm_glia$getDepth()
crm_glia$plotEmbedding(depth = TRUE, 
             depth.cutoff = 1e3)
```


```{r}
depth <- getConosDepth(crm_glia$con)

ggplot(data.frame(depth) %>% setNames(c("x")), aes(x)) + 
  geom_histogram(binwidth = 0.005) +
  xlim(c(0,1e6)) +
  scale_x_log10() +
  geom_vline(xintercept = 0.6e3, color="red") +
  labs(title="Depth Profile") +
  xlab("Cell size (UMIs)") +
  ylab("Number of cells")

filename = "figures/glia/depth_profile_glia_pre_filter.pdf"
pdf.width = 10
pdf.height = 8
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

```{r}
depth <- getConosDepth(crm_glia$con)
sum(depth<600)
```
```{r}
crm_glia$cms %>% sapply(ncol) %>% sum()
```

## Sample per cell 
```{r fig.height=8, fig.width=10}
dspc <- crm_glia$con$getDatasetPerCell()
```
```{r fig.height=8, fig.width=10}
crm_glia$con$plotGraph(groups=dspc,
                       show.legend=T,
                       size=0.1,
                       alpha=0.5,
                       font.size=0) %>% rasterise(dpi=600) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_blank())

filename = "figures/glia/dspc_glia_pre_filter.pdf"
pdf.width = 11.5
pdf.height = 10
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```


## Doublet detection 
```{r eval=FALSE, include=FALSE}
library(reticulate)
conda_create("r-reticulate", 
             conda = "/opt/software/miniconda/4.12.0/condabin/conda", 
             python_version = 3.8)
conda_install("r-reticulate", 
              conda = "/opt/software/miniconda/4.12.0/condabin/conda", 
              pip = TRUE, 
              packages = c("scrublet","doubletdetection"))
```
## Scrublet
```{r}
crm_glia$detectDoublets(conda.path = "/opt/software/miniconda/4.12.0/condabin/conda",method = "scrublet",env = "r-reticulate", verbose = T)
#Detected 147 possible doublets out of 27208 cells.
```

```{r fig.height=8, fig.width=8}
crm_glia$plotEmbedding(doublet.method = "scrublet")+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_blank())

filename = "figures/glia/scrublet_labels_glia.pdf"
pdf.width = 10
pdf.height = 10.2
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)

```
```{r fig.height=8, fig.width=8}
crm_glia$plotEmbedding(doublet.method = "scrublet", 
             doublet.scores = TRUE) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(),
          panel.border = element_blank())

filename = "figures/glia/scrublet_scores_glia.pdf"
pdf.width = 10
pdf.height = 10.2
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

## Mitochondrial gene fraction 

```{r fig.height=8, fig.width=8}
crm_glia$plotEmbedding(mito.frac = TRUE, 
             mito.cutoff = 0.05, 
             species = "human") +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(),
          panel.border = element_blank())

filename = "figures/glia/mitofraction_labels_glia.pdf"
pdf.width = 10
pdf.height = 10.2
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```
```{r fig.height=6, fig.width=10}
group.per.cell <- dspc %>% as.character()
mito_frac <- mitoFraction(crm_glia$con, "human")

ggplot(data.frame(Mito=mito_frac, Group=group.per.cell), aes(Group, Mito)) +
  ggbeeswarm::geom_quasirandom(size=0.1) +
  geom_hline(yintercept = 0.05, col="red") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ylab("Mitochondrial gene fraction") +
  xlab("Sample") +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_blank())

filename = "figures/glia/mitofraction_sample_scatter_glia.pdf"
pdf.width = 10
pdf.height = 8
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

## Filter plots
```{r fig.height=8, fig.width=10}
crm <- crm_glia

crm$plotFilteredCells(type="embedding",
                      doublet.method = "scrublet", 
                      depth = T,
                      depth.cutoff = 600,
                      mito.frac = TRUE, 
                      mito.cutoff = 0.05, 
                      species = "human",
                      cols = c("azure2",
                               "deeppink3",
                               "darkblue",
                               "orange",
                               "tan2",
                               "darkcyan",
                               "black",
                               "chartreuse1")
                      ,alpha=0.5) %>% rasterise(dpi=600) + 
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_blank())

filename = "figures/glia/cellstofilter_glia_prefilter_embedding.pdf"
pdf.width = 11.5
pdf.height = 10.2
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

```{r}
doublet_res <- crm_glia$doublets$scrublet$result
mito_res <- crm_glia$getMitoFraction()
crm_glia$cms %>% sapply(ncol) %>% sum
```
```{r}
filterVector <- CRMetrics:::filterVector
```

```{r}
library(ggrepel)

depth.cutoff=600
sep="!!"
depths <- crm_glia$getDepth() %>% filterVector("depth.cutoff", 
            depth.cutoff, depth.cutoff %>% names(), sep) %>% 
            {
                ifelse(!., "depth", "")
            }
    
mito.cutoff=0.05
species="human"
mf <- crm_glia$getMitoFraction(species = species) %>% filterVector("mito.cutoff", 
            mito.cutoff, mito.cutoff %>% names(), sep) %>% {
            ifelse(., "mito", "")
        }

doublet.method <- "scrublet"
tmp.doublets <- crm_glia$doublets[[doublet.method]]$result
        doublets <- tmp.doublets$labels %>% ifelse("doublet", 
            "") %>% setNames(rownames(tmp.doublets))

cell.idx <- crm_glia$cms %>% lapply(colnames) %>% Reduce(c, .)
tmp <- list(depth = depths, mito = mf, doublets = doublets) %>% 
        .[!sapply(., is.null)] %>% lapply(function(filter) filter[cell.idx]) %>% 
        bind_cols() %>% as.data.frame() %>% `rownames<-`(cell.idx)
    
tmp %<>% mutate(., filter = apply(., 1, paste, collapse = " ")) %>% 
            mutate(filter = gsub("^\\s+|\\s+$", "", filter) %>% 
                gsub("  ", " ", ., fixed = TRUE) %>% gsub(" ", 
                "+", .))
tmp$filter[tmp$filter == ""] <- "kept"
tmp$filter %<>% factor()
if ("kept" %in% levels(tmp$filter)) {
  tmp$filter %<>% relevel(ref = "kept")
  colstart <- 1
}

g <- tmp %>% mutate(., sample = rownames(.) %>% strsplit(sep) %>% 
                      sapply("[[", 1), filter = ifelse(grepl("+", filter, 
                                                             fixed = TRUE), "combination", as.character(filter))) %>% 
  group_by(sample, filter) %>% dplyr::count() %>% ungroup() %>% 
  group_by(sample) %>% mutate(pct = n/sum(n) * 100) %>% 
  ungroup() %>% filter(filter != "kept") %>% ggplot(aes(sample, 
                                                        pct, fill = filter)) + geom_bar(stat = "identity") + 
  geom_text_repel(aes(label = sprintf("%0.2f", round(pct, 
                                                     digits = 2))), position = position_stack(vjust = 0.5), 
                  direction = "y", size = 2.5) + crm$theme + theme(axis.text.x = element_text(angle = 45, 
                                                                                               hjust = 1)) + labs(x = "", y = "Percentage cells filtered")
g + theme(panel.border = element_blank())
filename = "figures/glia/cellstofilter_glia_prefilter_barplot.pdf"
pdf.width = 11.5
pdf.height = 10
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```


```{r}
crm_glia$filterCms(mito.cutoff =0.05,
                   species ="human",
                   doublets="scrublet",
                   depth.cutoff = 600)
```


```{r}
saveRDS(crm_glia, "data/crm_glia_filtered_manuscript.rds", compress=F)
```

# Neurons

```{r}
crm_neurons<- CRMetrics$new(cms = cms_neurons_clean, n.cores = 100)
```

```{r}
crm_neurons$addDetailedMetrics(n.cores = 100)
```

```{r}
dir.create("figures/neurons")
```

```{r fig.height=4, fig.width=8}
metrics.to.plot <- crm_neurons$detailed.metrics$metric %>%
  unique()
crm_neurons$plotDetailedMetrics(metrics = metrics.to.plot, 
                        plot.geom = "violin") +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank(), panel.border = element_blank())
filename = "figures/neurons/detailed_metrics_neurons.pdf"
pdf.width = 10
pdf.height = 5
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)

```
```{r}
crm_neurons$doPreprocessing(n.cores = 100)
```
```{r}
crm_neurons$createEmbedding(n.cores = 100)
```

```{r}
saveRDS(crm_neurons, "data/crm_neurons_cb_manuscript.rds", compress=F)
```


```{r}
crm_neurons$con$plotGraph(size=0.1,
                          shuffle.colors=T,
                          alpha=0.5) %>% rasterise(dpi=600) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(),
          panel.border = element_blank())

filename = "figures/neurons/UMAP_neurons.pdf"
pdf.width = 10
pdf.height = 10
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)

```

## General embedding
```{r fig.height=8, fig.width=8}
depth <- crm_neurons$getDepth()
crm_neurons$con$plotGraph(size=0.2, mark.groups=F, show.legend = T)


```
```{r fig.height=12, fig.width=9}
c("RBFOX3","SLC17A7","CUX2","RORB","FEZF2","THEMIS","GAD1","ID2","VIP","SST","PVALB") %>%
  lapply(function(g) crm_neurons$con$plotGraph(gene=g,
                                               title=g,
                                               size=0.03,
                                               alpha=0.3) %>% rasterise(dpi=600) + 
           theme(panel.grid.minor = element_blank(),
                 panel.grid.major = element_blank(),
                 panel.border = element_blank())) %>% 
  rasterize(dpi=450) %>%
  cowplot::plot_grid(plotlist=.,ncol=4)

filename = "figures/neurons/neurons_marker_genes.pdf"
pdf.width = 12
pdf.height = 9
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
  
```
## Depth filtering 

```{r}
depth <- getConosDepth(crm_neurons$con)

ggplot(data.frame(depth) %>% setNames(c("x")), aes(x)) +
  geom_histogram(binwidth = 0.005) + 
  xlim(c(0,1e6)) + scale_x_log10() + 
  geom_vline(xintercept = 1.7e3, color="red") + 
  labs(title="Depth Profile") + 
  xlab("Cell size (UMIs)") +
  ylab("Number of cells")

filename = "figures/neurons/depth_profile_neurons_pre_filter.pdf"
pdf.width = 10
pdf.height = 8
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

```{r}
depth <- getConosDepth(crm_neurons$con)
sum(depth<1700)
```
```{r}
crm_neurons$cms %>% sapply(ncol) %>% sum()
```

## Sample per cell 
```{r fig.height=8, fig.width=10}
dspc <- crm_neurons$con$getDatasetPerCell()
```
```{r fig.height=8, fig.width=10}
crm_neurons$con$plotGraph(groups=dspc, show.legend=T, size=0.1, alpha=0.5,font.size=0) %>% rasterise(dpi=600) +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank(), panel.border = element_blank())
filename = "figures/neurons/dspc_neurons_pre_filter.pdf"
pdf.width = 11.5
pdf.height = 10
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```


## Doublet detection (Not used)
```{r eval=FALSE, include=FALSE}
library(reticulate)
conda_create("r-reticulate", 
             conda = "/opt/software/miniconda/4.12.0/condabin/conda", 
             python_version = 3.8)
conda_install("r-reticulate", 
              conda = "/opt/software/miniconda/4.12.0/condabin/conda", 
              pip = TRUE, 
              packages = c("scrublet","doubletdetection"))
```
## Scrublet
```{r}
crm_neurons$detectDoublets(conda.path = "/opt/software/miniconda/4.12.0/condabin/conda",method = "scrublet",env = "r-reticulate", verbose = T)
#Detected 144 possible doublets out of 88407 cells.
```

```{r fig.height=8, fig.width=8}
crm_neurons$plotEmbedding(doublet.method = "scrublet") %>% rasterise(dpi=600)+
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank(), panel.border = element_blank())
filename = "figures/neurons/scrublet_labels_neurons.pdf"
pdf.width = 10
pdf.height = 10.2
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)

```
```{r fig.height=8, fig.width=8}
crm_neurons$plotEmbedding(doublet.method = "scrublet", 
             doublet.scores = TRUE) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(),
          panel.border = element_blank())

filename = "figures/neurons/scrublet_scores_neurons.pdf"
pdf.width = 10
pdf.height = 10.2
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

## Mitochondrial gene fraction 

```{r fig.height=8, fig.width=8}
crm_neurons$plotEmbedding(mito.frac = TRUE, 
             mito.cutoff = 0.05, 
             species = "human") +
    theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank(), panel.border = element_blank())
filename = "figures/neurons/mitofraction_labels_neurons.pdf"
pdf.width = 10
pdf.height = 10.2
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```
```{r fig.height=6, fig.width=10}
group.per.cell <- dspc %>% as.character()
mito_frac <- mitoFraction(crm_neurons$con, "human")
ggplot(data.frame(Mito=mito_frac, Group=group.per.cell), aes(Group, Mito)) + ggbeeswarm::geom_quasirandom(size=0.1) +
  geom_hline(yintercept = 0.05, col="red")+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  ylab("Mitochondrial gene fraction") +
  xlab("Sample") +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank(), panel.border = element_blank())
filename = "figures/neurons/mitofraction_sample_scatter_neurons.pdf"
pdf.width = 10
pdf.height = 8
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

## Filter plots
```{r fig.height=8, fig.width=10}
crm <- crm_neurons
crm_neurons$plotFilteredCells(type="embedding",
                      doublet.method = "scrublet", 
                      depth = T,
                      depth.cutoff = 1700,
                      mito.frac = TRUE, 
                      mito.cutoff = 0.05, 
                      species = "human",
                      cols = c("azure2","deeppink3","darkblue","orange","tan2","darkcyan","black","chartreuse1"),alpha=0.3)%>% rasterise(dpi=600) +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank(), panel.border = element_blank())
filename = "figures/neurons/cellstofilter_neurons_prefilter_embedding.pdf"
pdf.width = 11.5
pdf.height = 10.2
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

```{r fig.height=8, fig.width=10}
crm <- crm_neurons
crm_neurons$plotFilteredCells(type="bar",
                      doublet.method = "scrublet", 
                      depth = T,
                      depth.cutoff = 1700,
                      mito.frac = TRUE, 
                      mito.cutoff = 0.05, 
                      species = "human") + theme(panel.border = element_blank())
filename = "figures/neurons/cellstofilter_neurons_prefilter_bar.pdf"
pdf.width = 11.5
pdf.height = 10
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```
```{r}
library(ggrepel)
depth.cutoff=600
sep="!!"
depths <- crm_neurons$getDepth() %>% filterVector("depth.cutoff", 
            depth.cutoff, depth.cutoff %>% names(), sep) %>% 
            {
                ifelse(!., "depth", "")
            }
    
mito.cutoff=0.05
species="human"
mf <- crm_neurons$getMitoFraction(species = species) %>% filterVector("mito.cutoff", 
            mito.cutoff, mito.cutoff %>% names(), sep) %>% {
            ifelse(., "mito", "")
        }

doublet.method <- "scrublet"
tmp.doublets <- crm_neurons$doublets[[doublet.method]]$result
        doublets <- tmp.doublets$labels %>% ifelse("doublet", 
            "") %>% setNames(rownames(tmp.doublets))

cell.idx <- crm_neurons$cms %>% lapply(colnames) %>% Reduce(c, .)
tmp <- list(depth = depths, mito = mf, doublets = doublets) %>% 
        .[!sapply(., is.null)] %>% lapply(function(filter) filter[cell.idx]) %>% 
        bind_cols() %>% as.data.frame() %>% `rownames<-`(cell.idx)
    
tmp %<>% mutate(., filter = apply(., 1, paste, collapse = " ")) %>% 
            mutate(filter = gsub("^\\s+|\\s+$", "", filter) %>% 
                gsub("  ", " ", ., fixed = TRUE) %>% gsub(" ", 
                "+", .))
tmp$filter[tmp$filter == ""] <- "kept"
tmp$filter %<>% factor()
if ("kept" %in% levels(tmp$filter)) {
  tmp$filter %<>% relevel(ref = "kept")
  colstart <- 1
}

g <- tmp %>% mutate(., sample = rownames(.) %>% strsplit(sep) %>% 
                      sapply("[[", 1), filter = ifelse(grepl("+", filter, 
                                                             fixed = TRUE), "combination", as.character(filter))) %>% 
  group_by(sample, filter) %>% dplyr::count() %>% ungroup() %>% 
  group_by(sample) %>% mutate(pct = n/sum(n) * 100) %>% 
  ungroup() %>% filter(filter != "kept") %>% ggplot(aes(sample, 
                                                        pct, fill = filter)) + geom_bar(stat = "identity") + 
  geom_text_repel(aes(label = sprintf("%0.2f", round(pct, 
                                                     digits = 2))), position = position_stack(vjust = 0.5), 
                  direction = "y", size = 2.5) + crm$theme + theme(axis.text.x = element_text(angle = 45, 
                                                                                               hjust = 1)) + labs(x = "", y = "Percentage cells filtered")
g + theme(panel.border = element_blank())
filename = "figures/neurons/cellstofilter_neurons_prefilter_barplot.pdf"
pdf.width = 11.5
pdf.height = 10
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```
```{r}
crm_neurons$filterCms(depth.cutoff = 1700,mito.cutoff = 0.05,doublets = "scrublet",species = "human")
```


```{r}
saveRDS(crm_neurons, "data/crm_neurons_filtered_manuscript.rds", compress=F)
```


# New Conos Objects
```{r}
cms_glia_filter <- crm_glia$cms.filtered
cms_neurons_filter <- crm_neurons$cms.filtered
```


```{r}
con_glia<- quickConos(crm_glia$cms.filtered,
                  names(crm_glia$cms.filtered), 
                  make.geneknn = F,
                  get.largevis = F,
                  n.odgenes = 3e3,
                  nPcs = 100,
                  min.cells.per.gene = 3, 
                  min.transcripts.per.cell = 200, 
                  ncomps = 40,
                  get.tsne = T,
                  n.cores.p2=100,
                  n.cores.con=100)
con_glia <- con_glia$con

saveRDS(object = con_glia, file = "data/con_glia_filtered_manuscript.rds", compress = F)
```

```{r}
con_neurons<- quickConos(crm_neurons$cms.filtered,
                  names(crm_neurons$cms.filtered), 
                  make.geneknn = F,
                  get.largevis = F,
                  n.odgenes = 3e3,
                  nPcs = 100,
                  min.cells.per.gene = 3, 
                  min.transcripts.per.cell = 200, 
                  ncomps = 40,
                  get.tsne = T,
                  n.cores.p2=100,
                  n.cores.con=100)
con_neurons <- con_neurons$con

saveRDS(object = con_neurons, file = "data/con_neurons_filtered_manuscript.rds", compress = F)
```
```{r fig.width=8, fig.height=8}
con_neurons$plotGraph(size=0.1, alpha=.5, groups=annotation$l4, font.size=3)
```


#New CMS for filtered cells
```{r}
cms_filtered_n<- cms_neurons_filter
cms_filtered_n%<>% lapply(function(x) x[,colnames(x) %in% names(annotation$l1[annotation$l1 %in% c("Inhibitory neurons","Excitatory neurons")])])
```
```{r}
annotation_g <- readRDS("../../cacoa/Full_dataset/cacioa_all_2.0/annotation_all.rds")
annotation_g <- annotation_g$l1[annotation_g$l1 %in% c("Microglia","Astrocytes","OPCs","Oligodendrocytes","diff","Immune cells", "Endothelial cells", "Pericytes")] %>% droplevels()
cms_filtered_g <- cms_glia_filter
cms_filtered_g %<>% lapply(function(x) x[,colnames(x) %in% names(annotation_g[annotation_g %in% c("Astrocytes","diff","Endothelial cells","Immune cells","Microglia","Oligodendrocytes","OPCs","Pericytes")])])
```

#Make annotations and append them
```{r}
annotation_cms <- c(as.character(annotation_g), as.character(annotation$l1)) %>% setNames(c(names(annotation_g), names(annotation$l1))) %>% factor() %>% droplevels()
```

```{r}
cms_combined <- mapply(function(x, y) sccore::mergeCountMatrices(c(x, y)), x = crm_neurons$cms.filtered, y=crm_glia$cms.filtered)
cms_combined %>% sapply(ncol) %>% sum()
```

```{r}
con_filtered<- quickConos(cms_combined,
                  names(cms_combined), 
                  make.geneknn = F,
                  get.largevis = F,
                  n.odgenes = 3e3,
                  nPcs = 100,
                  min.cells.per.gene = 3, 
                  min.transcripts.per.cell = 200, 
                  ncomps = 40,
                  get.tsne = F,
                  n.cores.p2=100,
                  n.cores.con=100,)

con_filtered <- con_filtered$con

saveRDS(object = con_filtered, file = "data/con_filtered_manuscript.rds", compress = F)
```

```{r}
cms_combined_clean <- sapply(cms_combined,
                            modify_colnames,
                            simplify = FALSE)
```

```{r}
sessionInfo()
```

