---
title: "snRNA-seq TSC2: Transcription factor inference"
output: html_notebook
author: "Frederik Nørby Friis Sørensen, University of Copenhagen"
date: "April 2026"
---

```{r}
## We load the required packages
library(Seurat)
library(decoupleR)
library(OmnipathR)
library(cacoa)

# Only needed for data handling and plotting
library(dplyr)
library(tibble)
library(tidyr)
library(patchwork)
library(ggplot2)
library(pheatmap)
library(data.table)
```


```{r}
dir.create("../figures")
dir.create("../figures/inference")
```

### Get Collectri TF interactions for human
```{r}
net <- decoupleR::get_collectri(organism='human', split_complexes=FALSE)
net
```


###Load Cacoa object and extract DE analysis results

```{r}
cao <- readRDS("../../5.neurons/cao_neurons.rds")
```
```{r}
de <- cao$test.results$de
de <- lapply(de, "[[", "res")
```

### Extract Z-scores 

```{r}
dehm <- lapply(de, function(x){ 
    data.table(x) %>% 
    dplyr::select(Gene, Z) %>% 
    dplyr::filter(!is.na(Z)) %>% 
    column_to_rownames(var = "Gene") %>%
    as.matrix()})
```

### Perfom ULM pipeline with decoupler

```{r}
acts <- lapply(dehm, function(x) {run_ulm(x, network=net, .source='source', .target='target',
                .mor='mor', minsize = 5)})
```

### Make genes rows and clean up

```{r}
acts_pre <- lapply(acts, function(x){
  x %>%
    as.data.frame() %>% 
    set_rownames(x$source)
})
acts_pre <- lapply(acts_pre, "[", 4:4)
```


```{r}
saveRDS(acts_pre, "tf_activities_collectri.rds", compress=F)
```
```{r}
acts_pre <- readRDS("tf_activities_collectri.rds")
```


#Save data
```{r}
TF_neurons <- rbindlist(lapply(acts, function(x) x), idcol = "CellType")
write.csv2(TF_neurons,file = "Transcription_factor_inference_neurons.csv")
```


###Extract top 10 +NES and -NES TFs for each cell type and make union for HM plot


```{r}
#Remove celltypes
acts_pre$TBD <- NULL
acts_pre$Neu_dev_THY1 <- NULL
acts_top10_up <- lapply(acts_pre, function(x){
  x%>%
    rownames_to_column(var = "GeneID") %>%
    dplyr::rename(NES = "score") %>%
    dplyr::top_n(10, wt = NES) %>%
    dplyr::arrange(NES) %>% 
    dplyr::mutate(GeneID = factor(GeneID))})


acts_top10_down <- lapply(acts_pre, function(x){
  x%>%
    rownames_to_column(var = "GeneID") %>%
    dplyr::rename(NES = "score") %>%
    dplyr::top_n(10, wt = -NES) %>%
    dplyr::arrange(NES) %>% 
    dplyr::mutate(GeneID = factor(GeneID))})

top10up <- rbindlist(acts_top10_up)$GeneID %>% unique %>% as.character
top10down <- rbindlist(acts_top10_down)$GeneID %>% unique %>% as.character

m <- do.call(plyr::rbind.fill.matrix, lapply(acts_pre, t))
rownames(m) <- names(acts_pre)
m <- t(m)
m <- m[c(top10up, top10down),] %>% unique
```

### Extended data fig. 21
```{r fig.height=15 , fig.width=10}
library(ggplot2)
library(reshape2)
library(scales)
library(RColorBrewer)
cell_types <- c("L2_CUX2_LAMP5_PDGFD","L2_3_CUX2_FREM3_NTNG1", "L3_CUX2_SPOCK3_PRSS12", 
                "L4_RORB_SCHLAP1_ARHGAP15", "L4_RORB_SCHLAP1_MET", "L4_RORB_SCHLAP1_MME", 
                "L5_6_FEZF2_LRRK1_PCP4", "L5_6_FEZF2_LRRK1_SEMA3E", "L5_6_FEZF2_TLE4_ABO", 
                "L5_6_FEZF2_TLE4_HTR2C", "L5_6_FEZF2_TLE4_SCUBE1", "L5_6_THEMIS_NR4A2_NTNG2", 
                "L5_6_THEMIS_SEMA3A", "ID2_CHRNA7_CNR1", "ID2_CHRNA7_NCKAP5", "ID2_CHRNA7_PAX6",
                "ID2_LAMP5_CRH", "ID2_LAMP5_MEIS2", "ID2_LAMP5_NMBR", "ID2_LAMP5_NOS1", 
                "VIP_ABI3BP", "VIP_CBLN1", "VIP_CRH", "VIP_NRG1", "VIP_SEMA3C", "VIP_TRPM3_SSTR1", "VIP_TYR",
                "PVALB_BTBD11", "PVALB_LGR5", "PVALB_MEPE", "PVALB_UNC5B",
                "SST_CALB1", "SST_ISOC_THSD7B", "SST_NOS1", "SST_NPY", "SST_STK32A", "SST_TAC3", "SST_TH"
                )
genes_to_plot <- c("PLAGL1", "HOXB13", "NFIB", "SMAD4", "NR0B1", "SOX17", "STAT1", "TFCP2", "NKX2-2", "SPI1", "RFX1", "STAT4", "NEUROG1", "CRX", "CIITA", "KLF5", "NFATC2", "SRY", "E2F3", "KLF11", "EN1", "PREB", "IRF1", "NR1H2", "SNAI1", "LMO2", "KLF10", "NR1D1", "TBX2", "NR3C1", "SIM2", "ZEB2", "MZF1", "HOXA13", "HDAC7", "BCL11B", "ELF1", "THRB", "TFE3", "LMX1B", "NFIX", "SMARCA1", "HOXB7", "EGR2", "FOXA1", "SMAD3", "ZFPM1", "EP300", "ETV3", "APEX1", "HDAC9", "HDAC3", "MYCN", "IKZF1", "VDR", "RARA", "RARG", "MEF2D", "TFAP2C", "CEBPA", "IRF3", "CREB5", "ERF", "HNF1B", "WT1", "NCOR2", "YAP1", "BRCA1", "OTX2", "IRF8", "FOXA2", "HOXA10", "ESR2", "ESR1", "SMAD1", "ZNF382", "GATA5", "FOXN1", "SREBF1", "NR1H3", "TP63", "KCNIP3", "RUNX1", "IRF2", "RXRB", "SREBF2", "IRX1", "KDM5D", "ZNF331", "HTATIP2","REST","MECP2","RXRA","E2F2","DNMT3A","NR1I2","GTF2IRD1","E4F1","KDM5C")



# Assuming m is your matrix
m_rescaled <- scale(m[rowSums(ifelse(is.na(m), 1, 0)) < 10, ]) %>% rescale(to=c(-3,3))

# Melt the matrix into long format for ggplot2
m_melted <- melt(m_rescaled, varnames = c("Row", "Column"), value.name = "Value")

m_melted$Column <- factor(m_melted$Column, levels = cell_types)

m_melted <- subset(m_melted, Row %in% genes_to_plot)

dashed_line_position <- length(grep("^L", cell_types))

ggplot(m_melted, aes(x = Column, y = Row, fill = Value)) +
  geom_tile() +
  scale_fill_viridis(name = "NES", na.value = "gray",) + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.text.y = element_text(size = 8))

ggplot(m_melted, aes(x = Column, y = Row, fill = Value)) +
  geom_tile() +
 scale_fill_distiller(palette = "PuOr", name = "NES", na.value = "grey")+
  theme_minimal() +
  geom_vline(xintercept = dashed_line_position + 0.5, linetype = "dashed", color = "black") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        axis.text.y = element_text(size = 8))+
  ylab(element_blank()) +
  xlab(element_blank())

filename = "../figures/inference/TF_Heatmap_Big.pdf"
pdf.width = 10
pdf.height = 15
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)

```

### Plot top genes
### Fig. 6c
```{r fig.height=5, fig.width=10, paged.print=FALSE}
library(ggplot2)
library(reshape2)
library(scales)
library(viridis)
library(dplyr)
library(data.table)
library(plyr)


cell_types <- c("L2_CUX2_LAMP5_PDGFD","L2_3_CUX2_FREM3_NTNG1", "L3_CUX2_SPOCK3_PRSS12", 
                "L4_RORB_SCHLAP1_ARHGAP15", "L4_RORB_SCHLAP1_MET", "L4_RORB_SCHLAP1_MME", 
                "L5_6_FEZF2_LRRK1_PCP4", "L5_6_FEZF2_LRRK1_SEMA3E", "L5_6_FEZF2_TLE4_ABO", 
                "L5_6_FEZF2_TLE4_HTR2C", "L5_6_FEZF2_TLE4_SCUBE1", "L5_6_THEMIS_NR4A2_NTNG2", 
                "L5_6_THEMIS_SEMA3A", "ID2_CHRNA7_CNR1", "ID2_CHRNA7_NCKAP5", "ID2_CHRNA7_PAX6",
                "ID2_LAMP5_CRH", "ID2_LAMP5_MEIS2", "ID2_LAMP5_NMBR", "ID2_LAMP5_NOS1", 
                "VIP_ABI3BP", "VIP_CBLN1", "VIP_CRH", "VIP_NRG1", "VIP_SEMA3C", "VIP_TRPM3_SSTR1", "VIP_TYR",
                "PVALB_BTBD11", "PVALB_LGR5", "PVALB_MEPE", "PVALB_UNC5B",
                "SST_CALB1", "SST_ISOC_THSD7B", "SST_NOS1", "SST_NPY", "SST_STK32A", "SST_TAC3", "SST_TH"
                )

# Define the genes to plot
downgenes <- c("MYC", "E2F4", "HIF1A", "NFKB", "CREB1", "JUN")
upgenes <- c("REST", "DNMT3A", "E2F2", "GTF2IRD1", "NR1I2", "ZNF331","CDX2","GLI3","SOX5","CREBBP","MECP2","ERF","KDM5C")
genes_to_plot <- c(downgenes, upgenes)

# Filter the matrix for the specified genes
m <- do.call(plyr::rbind.fill.matrix, lapply(acts_pre, t))
rownames(m) <- names(acts_pre)
m <- t(m)

# Check which genes are present in the matrix
present_genes <- genes_to_plot[genes_to_plot %in% rownames(m)]
missing_genes <- genes_to_plot[!genes_to_plot %in% rownames(m)]

# Print missing genes
if (length(missing_genes) > 0) {
  cat("The following genes are missing from the matrix:\n")
  print(missing_genes)
}

# Proceed with the present genes
m_filtered <- m[present_genes, , drop = FALSE]

# Rescale the matrix
m_rescaled <- scale(m_filtered[rowSums(ifelse(is.na(m_filtered), 1, 0)) < 10, ]) %>% rescale(to = c(-3, 3))

# Melt the matrix into long format for ggplot2
m_melted <- melt(m_rescaled, varnames = c("Row", "Column"), value.name = "Value")
# Ensure the x-axis follows the desired order
m_melted$Column <- factor(m_melted$Column, levels = cell_types)

#Remove NA cell types
m_melted<- m_melted %>% drop_na(Column)

# Determine the position for the dashed line
dashed_line_position <- length(grep("^L", cell_types))

# Plot using the Viridis color scale
ggplot(m_melted, aes(x = Column, y = Row, fill = Value)) +
  geom_tile() +
  scale_fill_viridis(name = "NES", na.value = "gray") +
  theme_minimal() +
  geom_vline(xintercept = dashed_line_position + 0.5, linetype = "dashed", color = "black") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        axis.text.y = element_text(size = 8))+
  ylab(element_blank())+
  xlab(element_blank())
filename = "../figures/inference/TF_up_down_viridis.pdf"
pdf.width = 10
pdf.height = 5
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)

  # Plot using the PuOr color scale
ggplot(m_melted, aes(x = Column, y = Row, fill = Value)) +
  geom_tile() +
  scale_fill_distiller(palette = "PuOr", name = "NES", na.value = "gray") +
  theme_minimal() +
  geom_vline(xintercept = dashed_line_position + 0.5, linetype = "dashed", color = "black") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        axis.text.y = element_text(size = 8))+
  ylab(element_blank())+
  xlab(element_blank())
filename = "../figures/inference/TF_up_down_PuOr.pdf"
pdf.width = 10
pdf.height = 5
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

# Pathway activity inference from scRNA-seq

```{r}
## We load the required packages
library(Seurat)
library(decoupleR)

# Only needed for data handling and plotting
library(dplyr)
library(tibble)
library(tidyr)
library(patchwork)
library(ggplot2)
library(pheatmap)
library(scales)
```

```{r}
net <- get_progeny(organism = 'human', top = 500)
net
```

```{r}
dehm <- lapply(de, function(x){ 
    data.table(x) %>% 
    dplyr::select(Gene, Z) %>% 
    dplyr::filter(!is.na(Z)) %>% 
    column_to_rownames(var = "Gene") %>%
    as.matrix()})
```


```{r}
acts <- lapply(dehm, function(x) {run_mlm(x, net=net, .source='source', .target='target',
                .mor='weight', minsize = 5)})
```

```{r}
acts$TBD <- NULL
acts$Neu_dev_THY1 <- NULL
combined_data <- do.call(rbind, lapply(names(acts), function(cell_type) {
  data <- acts[[cell_type]]
  data$cell_type <- cell_type
  return(data)
}))

p_value_to_asterisks <- function(p) {
  if (p < 0.001) {
    return("***")
  } else if (p < 0.01) {
    return("**")
  } else if (p < 0.05) {
    return("*")
  } else {
    return("")
  }
}

combined_data <- combined_data %>%
  mutate(asterisks = sapply(p_value, p_value_to_asterisks))

combined_data$scaled_score <- scale(combined_data$score) %>% rescale(to = c(-3, 3))
names(combined_data)
```
## Pathway activity inference plot
### Fig. 6d

```{r}
cell_types <- c("L2_CUX2_LAMP5_PDGFD","L2_3_CUX2_FREM3_NTNG1", "L3_CUX2_SPOCK3_PRSS12", 
                "L4_RORB_SCHLAP1_ARHGAP15", "L4_RORB_SCHLAP1_MET", "L4_RORB_SCHLAP1_MME", 
                "L5_6_FEZF2_LRRK1_PCP4", "L5_6_FEZF2_LRRK1_SEMA3E", "L5_6_FEZF2_TLE4_ABO", 
                "L5_6_FEZF2_TLE4_HTR2C", "L5_6_FEZF2_TLE4_SCUBE1", "L5_6_THEMIS_NR4A2_NTNG2", 
                "L5_6_THEMIS_SEMA3A", "ID2_CHRNA7_CNR1", "ID2_CHRNA7_NCKAP5", "ID2_CHRNA7_PAX6",
                "ID2_LAMP5_CRH", "ID2_LAMP5_MEIS2", "ID2_LAMP5_NMBR", "ID2_LAMP5_NOS1", 
                "VIP_ABI3BP", "VIP_CBLN1", "VIP_CRH", "VIP_NRG1", "VIP_SEMA3C", "VIP_TRPM3_SSTR1", "VIP_TYR",
                "PVALB_BTBD11", "PVALB_LGR5", "PVALB_MEPE", "PVALB_UNC5B",
                "SST_CALB1", "SST_ISOC_THSD7B", "SST_NOS1", "SST_NPY", "SST_STK32A", "SST_TAC3", "SST_TH"
                )

combined_data$cell_type <- factor(combined_data$cell_type, levels = cell_types)
ggplot(combined_data, aes(x = cell_type, y = source, fill = scaled_score)) +
  geom_tile() +
  geom_text(aes(label = asterisks), color = "white", size = 3) +
  scale_fill_distiller(palette = "PuOr", name = "NES", na.value = "gray") +
  theme_minimal() +
  geom_vline(xintercept = dashed_line_position + 0.5, linetype = "dashed", color = "black") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        axis.text.y = element_text(size = 8))+
  ylab(element_blank())+
  xlab(element_blank())
filename = "../figures/inference/Pathway_activity_inference_PuOr.pdf"
pdf.width = 10
pdf.height = 5
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

```{r}
sessionInfo()
```


