---
title: "snRNA-seq TSC2 preprocessing without Cellbender"
output: html_notebook
author: "Frederik Nørby Friis Sørensen, University of Copenhagen"
date: "April 2026"
---

### Setup libraries
```{r setup}
library(dplyr)
library(magrittr)
library(conos)
library(pagoda2)
library(cacoa)
library(reticulate)
library(ggplot2)
library(ggrastr)
library(cowplot)
library(HDF5Array)
library(Seurat)
library(CRMetrics)
library(scHelper)
set.seed(1)
```

```{r message=FALSE, include=FALSE}
# Check if 'data' folder exists, if not, create it
if (!dir.exists("data")) {
  dir.create("data")
  cat("Created 'data' folder\n")
} else {
  cat("'data' folder already exists\n")
}

# Check if 'figure' folder exists, if not, create it
if (!dir.exists("figures")) {
  dir.create("figures")
  cat("Created 'figures' folder\n")
} else {
  cat("'figures' folder already exists\n")
}

```

# CRMetrics
## Make object and load count matrices
### Replace data path with the location you store the count matrices. 
```{r}
data_path= "pathtocountmatrices"
crm <- CRMetrics$new(data.path = data_path, n.cores = 100)
crm$addCms()
```

```{r}
crm$selectMetrics()
```

```{r fig.height=12, fig.width=12}
metrics.to.plot <- crm$selectMetrics(ids = c(1:4,6,18,19,15,9))
crm$plotSummaryMetrics(metrics = metrics.to.plot, 
                       plot.geom = "bar")
filename = "figures/Metrics_cellranger.pdf"
pdf.width = 12
pdf.height = 12
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```


```{r}
crm$addDetailedMetrics(n.cores = 100)
```

```{r fig.height=4, fig.width=8}
metrics.to.plot <- crm$detailed.metrics$metric %>%
  unique()
crm$plotDetailedMetrics(metrics = metrics.to.plot, 
                        plot.geom = "violin")
```
## Preprocess count matrices with pagoda2 and make initial conos object:

```{r include=FALSE}
crm$doPreprocessing(n.cores = 100)
```
```{r include=FALSE}
crm$createEmbedding(n.cores = 100)
```


```{r fig.height=20, fig.width=15}
c("RBFOX3","GAD1","SLC17A7","VIP","ID2","SST","PVALB","MOG","VCAN","CSF1R","AQP4","CLDN5") %>%
  lapply(function(g) crm$con$plotGraph(gene=g, title=g, size=0.1, alpha=1) %>% rasterize(dpi=600) + 
           theme_void() + 
           theme(legend.position = "none")) %>%
  cowplot::plot_grid(plotlist=., ncol=3) %>% rasterize(dpi=300)
filename = "figures/Markers_pre_filter.pdf"
pdf.width = 15
pdf.height = 20
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

```{r}
joint <- crm$con$getJointCountMatrix()
```

```{r}
native <- getConosCluster(crm$con)
```

```{r fig.height=6, fig.width=9 }
markers<- c("RBFOX3","GAD1","SLC17A7","VIP","ID2","SST","PVALB","MOG","VCAN","CSF1R","AQP4","CLDN5")
sccore:::dotPlot(markers, joint, cell.groups = native, gene.order = markers,col.max = 1.5)+scale_color_viridis_c()
filename = "figures/dotplot_markers_prefilter.pdf"
pdf.width = 9
pdf.height = 6
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

## General embedding
```{r fig.height=8, fig.width=10}
crm$con$plotGraph(size=0.1, shuffle.colors=T, alpha=0.2, font.size=3) %>% rasterize(dpi=600) +
  theme_void()
filename = "figures/UMAP_prefilter.pdf"
pdf.width = 10
pdf.height = 8
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```
## Depth filtering 

```{r fig.height=8, fig.width=8}
p <- crm$plotEmbedding(depth = TRUE, depth.cutoff = 750) +
  theme_void() +
  theme(legend.position = "none") +
  xlab("UMAP1") +
  ylab("UMAP2") 
ggsave(plot = p, filename = "figures/UMAP_depth_pre_filter.pdf",dpi=300,width = 10, height = 8,)
p
```


## Doublet detection 
```{r eval=FALSE, include=FALSE}
library(reticulate)
conda_create("r-reticulate", 
             conda = "/opt/software/miniconda/4.12.0/condabin/conda", 
             python_version = 3.8)
conda_install("r-reticulate", 
              conda = "/opt/software/miniconda/4.12.0/condabin/conda", 
              pip = TRUE, 
              packages = c("scrublet","doubletdetection"))
```
### Scrublet
```{r}
crm$detectDoublets(conda.path = "/opt/software/miniconda/4.12.0/condabin/conda", method = "scrublet",env = "r-reticulate", verbose = T)
```

```{r fig.height=8, fig.width=8}
crm$plotEmbedding(doublet.method = "scrublet") +
  theme_minimal() +
  xlab("UMAP1") +
  ylab("UMAP2") +
  theme(legend.position = "none")

filename = "figures/UMAP_scrublet.pdf"
pdf.width = 9
pdf.height = 9
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

```{r fig.height=8, fig.width=8}
crm$plotEmbedding(doublet.method = "scrublet", 
             doublet.scores = TRUE) +
  theme_minimal() +
  xlab("UMAP 1") +
  ylab("UMAP 2")
```

## Mitochondrial gene fraction 

```{r fig.height=8, fig.width=8}
crm$plotEmbedding(mito.frac = TRUE, 
             mito.cutoff = 0.05, 
             species = "human") +
  theme_minimal() +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  theme(legend.position = "none")

filename = "figures/UMAP_mitofrac.pdf"
pdf.width = 9
pdf.height = 9
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```
```{r fig.height=6, fig.width=10}
dspc <- crm$con$getDatasetPerCell()
group.per.cell <- dspc %>% as.character()

mito_frac <- mitoFraction(crm$con, "human")

ggplot(data.frame(Mito=mito_frac, Group=group.per.cell), aes(Group, Mito)) + ggbeeswarm::geom_quasirandom(size=0.1) %>% rasterize(dpi=600)+
  geom_hline(yintercept = 0.05, col="red")+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  ylab("Mitochondrial gene fraction") +
  xlab("Sample") +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank(), panel.border = element_blank())


filename = "figures/mitofraction_sample_scatter_all_cells.pdf"
pdf.width = 10
pdf.height = 6
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```
## Scrublet filter plots
```{r fig.height=8, fig.width=10}
crm$plotFilteredCells(type="embedding",
                      depth = TRUE, 
                      depth.cutoff = 750, 
                      doublet.method = "scrublet", 
                      mito.frac = TRUE, 
                      mito.cutoff = 0.05, 
                      species = "human",
                      cols = c("aliceblue","deeppink3","darkblue","orange","tan2","darkcyan","black","chartreuse1")) %>% rasterize(dpi=600) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),panel.grid = element_blank(),panel.background = element_blank(),panel.grid.minor = element_blank(),axis.title.x =
           element_blank(),axis.title.y = element_blank(),axis.ticks.x = element_blank(),axis.ticks.y = element_blank(),axis.text.x  = element_blank(),axis.text.y  = element_blank())

filename = "figures/UMAP_filtered_cells.pdf"
pdf.width = 11
pdf.height = 9
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

```{r fig.height=8, fig.width=10}
crm$plotFilteredCells(type = "bar", 
                      doublet_method = "scrublet", 
                      depth = TRUE, 
                      depth.cutoff = 750, 
                      mito.frac = TRUE, 
                      mito.cutoff = 0.05, 
                      species = "human")
filename = "figures/bar_plot_filtered_cells.pdf"
pdf.width = 11
pdf.height = 9
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

```{r}
saveRDS(crm, "data/crm_ready_filtering_no_cb.rds", compress=F)
```

## Filter and save count matrices
```{r}
crm$filterCms(depth.cutoff = 0.75e3,
              mito.cutoff =0.05,
              species ="human",
              doublets="scrublet")

saveRDS(crm$cms.filtered, "data/cms_filtered_no_cb_750.rds", compress=F)
```


## Load CMS to make new conos object
```{r}
cms <- readRDS("data/cms_filtered_no_cb_750.rds")
```
```{r}
print(length(cms))
print(length(names(cms)))
print(any(is.na(names(cms))))
print(sapply(cms, function(x) any(duplicated(colnames(x)))))
length(cms) == length(names(cms))
```

## Make Conos object
```{r eval=FALSE, include=TRUE}
con<- quickConos(cms = cms,
                 sample.names = names(cms), 
                 make.geneknn = F,
                 get.largevis = F,
                 n.odgenes = 3e3,
                 nPcs = 100,
                 min.cells.per.gene = 3, 
                 min.transcripts.per.cell = 200, 
                 ncomps = 40,
                 get.tsne = F,
                 n.cores.p2=100,
                 n.cores.con=100)
con <- con$con

saveRDS(object = con, file = "data/con_filtered_no_cb_750.rds", compress = F)
```
```{r eval=FALSE, include=FALSE}
saveRDS(crm, "data/crm_filtered_no_cb.rds", compress=F)
```

```{r fig.width=9, fig.height=9}
con$plotGraph(shuffle.colors=T, alpha=0.5, size=0.1, groups = annotation$anno_mid)
filename = "figures/UMAP_filtered.pdf"
pdf.width = 9
pdf.height = 9
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

```{r}
sessionInfo()
```

