---
title: "RNAScope excitatory marker validation"
output: html_notebook
author: "Frederik Nørby Friis Sørensen, University of Copenhagen"
date: "April 2026"
---


```{r LIBS}
library(ggplot2)
library(tidyr)
library(dplyr)
library(ggpubr)
library(ARTool)
library(emmeans)
library(readr)
```


## Relative densities (no. of  positive cells in bin / total positive cells in cortical column per gene between conditions (Analysis of Variance of Aligned Rank Transformed Data):
# Fig 7g
```{r fig.height=5.5, fig.width=3}
# Load data
data <- read.csv("RNAscope_data_for_R.csv", header = TRUE)

genes_of_interest <- c("CUX2", "RORB", "FEZF2")

# Prepare data
long_data <- data %>%
  mutate(Condition = ifelse(grepl("control", Sample, ignore.case = TRUE), "Control", "TSC")) %>%
  pivot_longer(cols = all_of(genes_of_interest), names_to = "Gene", values_to = "RawCount") %>%
  group_by(Sample, Gene) %>%
  mutate(TotalCount = sum(RawCount[Bin %in% 1:7], na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    RelativeCount = RawCount / TotalCount,
    Gene = factor(Gene, levels = genes_of_interest),
    Bin = factor(Bin, levels = as.character(1:7)),
    Condition = factor(Condition)
  )

long_data$Gene <- factor(long_data$Gene,
                            levels = genes_of_interest)

# --- 1) Run ART per gene and extract overall Condition p-value ---
art_results <- lapply(genes_of_interest, function(gene_name) {
  df <- long_data %>% filter(Gene == gene_name, !is.na(RelativeCount))
  art_mod <- art(RelativeCount ~ Condition * Bin, data = df)
  anova_res <- anova(art_mod)
  
  # Overall Condition p-value
  cond_p <- anova_res["Condition", "Pr(>F)"]
  
  list(gene = gene_name, art_mod = art_mod, condition_p = cond_p)
})

overall_pvals <- data.frame(
  Gene = sapply(art_results, `[[`, "gene"),
  Condition_overall_p = sapply(art_results, `[[`, "condition_p"),
  p_label = paste0("p=", round(sapply(art_results, `[[`, "condition_p"), 3))
)

# --- 2) Posthoc: pairwise per Bin ---
posthoc_list <- lapply(art_results, function(res) {
  df <- long_data %>% filter(Gene == res$gene, !is.na(RelativeCount))
  
  # Extract lm object for interaction term
  art_lm <- artlm(res$art_mod, "Condition:Bin")
  
  # Compute estimated marginal means
  emm <- emmeans(art_lm, ~ Condition | Bin)
  pw <- contrast(emm, method = "pairwise")
  pw_df <- as.data.frame(pw)
  
  # Bonferroni correction
  pw_df$p.adj <- p.adjust(pw_df$p.value, method = "bonferroni")
  pw_df <- pw_df %>% mutate(
    signif = ifelse(p.adj < 0.05, TRUE, FALSE),
    Gene = res$gene
  )
  pw_df
})

posthoc_df <- do.call(rbind, posthoc_list)
overall_pvals$Gene <- factor(overall_pvals$Gene, levels = genes_of_interest)
posthoc_df$Gene <- factor(posthoc_df$Gene, levels = genes_of_interest)

# --- 3) Plot ---
gg_panelH <- ggplot(long_data, aes(x = Bin, y = RelativeCount, color = Condition)) +
  geom_jitter(aes(shape = Condition), width = 0.1, size = 1.5, alpha = 0.9) +
  stat_summary(fun = mean, geom = "line", aes(group = Condition, linetype = Condition), size = 0.9) +
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(group = Condition), width = 0.3) +
  geom_text(
    data = posthoc_df %>% filter(signif == TRUE),
    aes(x = Bin, y = 1.05, label = "*"), # 1.05 is relative; adjust per gene if needed
    inherit.aes = FALSE,
    size = 4
  ) +
  geom_text(
    data = overall_pvals,
    aes(x = 6, y = 1.1, label = p_label),
    inherit.aes = FALSE,
    size = 3.5
  ) +
  facet_wrap(~Gene, scales = "free_y", ncol = 1, drop=FALSE) +
  scale_color_manual(values = c("Control" = "dodgerblue3", "TSC" = "deeppink3")) +
  scale_shape_manual(values = c("Control" = 16, "TSC" = 17)) +
  scale_linetype_manual(values = c("Control" = "solid", "TSC" = "solid")) +
  labs(x = "Bin", y = "Relative density") +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "bottom",
    legend.title = element_blank()
  )

gg_panelH


ggsave(plot = gg_panelH,
       filename = "figures/RNAscope_relative_abundance_lines_new.pdf",
       dpi = 600,
       height = 5.5,
       width = 3.5)

write_csv(posthoc_df, "RNAscope_Statistics_CUX2_RORB_FEZF2_relative_abundance_posthoc.csv")

```


## Boxplots for total cell density over cortical column
# Fig 7h
```{r fig.height=5, fig.width=3.5}
# Required libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(exactRankTests)

# Load data
data <- read.csv("RNAscope_data_for_R.csv",header = T)

# Annotate condition
data <- data %>%
  mutate(Condition = ifelse(grepl("control", Sample, ignore.case = TRUE), "Control", "TSC")) 

# Markers
genes <- c("CUX2", "RORB", "FEZF2", "CUX2.RORB", "RORB.FEZF2", "CUX2.FEZF2")
area_mm2 <- 0.4138*7 # Bin-size in mm^2 * total number of bins

# Calculate densities
density_data <- data %>%
  #filter(Sample != "tsc-04") %>%
  filter(Bin %in% 1:7) %>%
  group_by(Sample, Condition) %>%
  summarise(across(all_of(genes), sum, na.rm = TRUE), .groups = "drop") %>%
  pivot_longer(cols = all_of(genes), names_to = "Gene", values_to = "PositiveCells") %>%
  mutate(Density = PositiveCells / area_mm2)

# Set factor levels
density_data$Gene <- factor(density_data$Gene,
                            levels = c("CUX2", "RORB", "FEZF2", "CUX2.RORB", "RORB.FEZF2", "CUX2.FEZF2"))

# Max y for text placement
y_max <- density_data %>%
  group_by(Gene) %>%
  summarise(y.position = max(Density) * 1.1, .groups = "drop")

# Wilcoxon exact test per gene
stats_table <- density_data %>%
  group_by(Gene) %>%
  summarise(
    p_value = wilcox.exact(Density ~ Condition, data = cur_data(), exact = TRUE)$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    p_adj = p.adjust(p_value, method = "fdr"),
    p_signif = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01 ~ "**",
      p_adj < 0.05 ~ "*",
      TRUE ~ "ns"
    )
  ) %>%
  left_join(y_max, by = "Gene") %>%
  mutate(x = 1.5)

stats_table <- stats_table %>%
  mutate(
    p_label = paste0((p_signif),"\np=", sprintf("%.3f", p_adj))
  )


# Plot
gg_density_boxplot <- ggplot(density_data, aes(x = Condition, y = Density, fill = Condition)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, width = 0.6) +
  geom_jitter(aes(shape = Condition), width = 0.15, size = 2, alpha = 0.7) +
 geom_text(
  data = stats_table,
  aes(x = x, y = y.position, label = p_label),
  size = 3.5,
  lineheight=0.8,
  vjust=-.5,
  inherit.aes = FALSE ) +
  facet_wrap(~Gene, ncol = 3, nrow = 2, scales = "free_y") +
  scale_fill_manual(values = c("Control" = "dodgerblue2", "TSC" = "deeppink3")) +
  scale_shape_manual(values = c("Control" = 16, "TSC" = 17)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.3))) +
  labs(
    x = NULL,
    y = expression("Cell density ("*cells/mm^2*")")
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 9),
    axis.text.x = element_blank()
  )

# View results
print(stats_table)
print(gg_density_boxplot)

ggsave(plot = gg_density_boxplot,
       filename = "figures/RNAscope_Boxplot_exci_Total_Positive_cell_density.pdf",
       dpi = 600,
       height = 5,
       width = 3.5)

write_csv(stats_table, "stats_table_RNAscope_Boxplot_exci_Total_Positive_cell_density.csv")


```
```{r}
# Calculate Wilcoxon test and adjusted p-values per gene
stats_table <- density_data %>%
  group_by(Gene) %>%
  summarise(
    p_value = exactRankTests:::wilcox.exact(Density ~ Condition)$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    p_adj = p.adjust(p_value, method = "fdr"),
    p_signif = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01 ~ "**",
      p_adj < 0.05 ~ "*",
      TRUE ~ "ns"
    )
  )

# View results
print(stats_table)

```

## CUX2 ratio across upper and lower layers (bins)
```{r fig.height=2.5, fig.width=1.5}
# Prepare ratio data, impute 1 for bin5_7 == 0
ratio_data <- data %>%
  filter(Bin %in% 1:7) %>%
  group_by(Sample, Condition) %>%
  summarise(
    bin1_3 = sum(CUX2[Bin %in% 1:3], na.rm = TRUE),
    bin5_7 = sum(CUX2[Bin %in% 5:7], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # Replace 0 in bin5_7 with 1
  mutate(bin5_7 = ifelse(bin5_7 == 0, 1, bin5_7)) %>%
  mutate(Ratio = bin1_3 / bin5_7)

# Compute Wilcoxon test
ratio_test <- wilcox.test(Ratio ~ Condition, data = ratio_data, exact = FALSE)
p_label_ratio <- paste0("p=", sprintf("%.2f", ratio_test$p.value))

# Plot as violin + boxplot
gg_ratio_violin <- ggplot(ratio_data, aes(x = Condition, y = Ratio, fill = Condition)) +
  geom_violin(trim = T, width = 0.6, alpha = 0.7) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.1, size = 1.6, alpha = 0.85) +
  annotate(
    "text",
    x = 1.5,
    y = max(ratio_data$Ratio, na.rm = TRUE) * 1.05,
    label = p_label_ratio,
    size = 3
  ) +
  scale_fill_manual(values = c("Control" = "dodgerblue2", "TSC" = "deeppink3")) +
  labs(y = "Bin ratio (1–3 / 5–7)",
       x = NULL,
       title = "CUX2") +
  theme_minimal() +
  theme(title = element_text(face="bold"),
    legend.position = "bottom",
    legend.title = element_blank(),
    strip.text = element_text(face = "bold", size = 9),
    axis.text.x = element_blank()
  )

gg_ratio_violin

ggsave(plot = gg_ratio_violin,
       filename = "figures/RNAscope_Boxplot_CUX2_ratio_bin1-3vs1-7.pdf",
       dpi = 600,
       height = 2.5,
       width = 1.5)
```

## Do comparison of summarized bins between condition:
```{r }
library(dplyr)
library(tidyr)

# Summarize per sample for upper and lower bins
ratio_data_long <- ratio_data %>%
  mutate(
    BinGroup = "Bin1-3",
    Value = bin1_3
  ) %>%
  dplyr::select(Sample, Condition, BinGroup, Value) %>%
  bind_rows(
    ratio_data %>%
      mutate(
        BinGroup = "Bin5-7",
        Value = bin5_7
      ) %>%
      dplyr::select(Sample, Condition, BinGroup, Value)
  ) %>%
  mutate(BinGroup = factor(BinGroup, levels = c("Bin1-3", "Bin5-7")))

```

# Fig 7i CUX2 panels
```{r fig.height=2.5, fig.width=2.5}
stats_ratio <- ratio_data_long %>%
  group_by(BinGroup) %>%
  summarise(
    p_val = wilcox.test(Value ~ Condition)$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    p_adj = p.adjust(p_val, method = "fdr"),   # FDR correction
    p_sign = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE ~ "ns"
    ),
    x = 1.5,  # midpoint between Control and TSC
    y.position = ratio_data_long %>%
      group_by(BinGroup) %>%
      summarise(max_val = max(Value, na.rm = TRUE)) %>%
      pull(max_val) * 1.05,
    p_label = paste0(p_sign, "\np=", sprintf("%.3f", p_adj))
   )


gg_ratio_box <- ggplot(ratio_data_long, aes(x = Condition, y = Value, fill = Condition, shape = Condition)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, width = 0.6) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
  geom_text(
    data = stats_ratio,
    aes(x = x, y = y.position, label = p_label),
    inherit.aes = FALSE,
    size = 3.5,
    lineheight = 0.8,
    vjust = -0.5
  ) +
  facet_wrap(~BinGroup, scales = "free_y") +
  scale_fill_manual(values = c("Control" = "dodgerblue2", "TSC" = "deeppink3")) +
  scale_shape_manual(values = c("Control" = 16, "TSC" = 17)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.3))) +
  labs(y = "CUX2 positive cells", x = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    strip.text = element_text(face = "bold", size = 9),
    axis.text.x = element_blank()
  )

#Plot
gg_ratio_box
ggsave(plot = gg_ratio_box,
       filename = "figures/RNAscope_Boxplot_CUX2_positive_cells_1-3+1-7_conditions.pdf",
       dpi = 600,
       height = 2.5,
       width = 2.5)

```


## And now for FEZF2
```{r fig.height=2.5, fig.width=1.5}
# Prepare ratio data, impute 1 for bin5_7 == 0
ratio_data <- data %>%
  filter(Bin %in% 1:7) %>%
  group_by(Sample, Condition) %>%
  summarise(
    bin1_3 = sum(FEZF2[Bin %in% 1:3], na.rm = TRUE),
    bin5_7 = sum(FEZF2[Bin %in% 5:7], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # Replace 0 in bin5_7 with 1
  mutate(bin5_7 = ifelse(bin5_7 == 0, 1, bin5_7)) %>%
  mutate(Ratio = bin1_3 / bin5_7)

# Compute Wilcoxon test
ratio_test <- wilcox.test(Ratio ~ Condition, data = ratio_data, exact = FALSE)
p_label_ratio <- paste0("p=", sprintf("%.2f", ratio_test$p.value))

# Plot as violin + boxplot
gg_ratio_violin <- ggplot(ratio_data, aes(x = Condition, y = Ratio, fill = Condition)) +
  geom_violin(trim = T, width = 0.6, alpha = 0.7) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.1, size = 1.6, alpha = 0.85) +
  annotate(
    "text",
    x = 1.5,
    y = max(ratio_data$Ratio, na.rm = TRUE) * 1.05,
    label = p_label_ratio,
    size = 3
  ) +
  scale_fill_manual(values = c("Control" = "dodgerblue2", "TSC" = "deeppink3")) +
  labs(y = "Bin ratio (1–3 / 5–7)",
       x = NULL,
       title = "FEZF2") +
  theme_minimal() +
  theme(title = element_text(face="bold"),
    legend.position = "bottom",
    legend.title = element_blank(),
    strip.text = element_text(face = "bold", size = 9),
    axis.text.x = element_blank()
  )

gg_ratio_violin

ggsave(plot = gg_ratio_violin,
       filename = "figures/RNAscope_Boxplot_FEZF2_ratio_bin1-3vs1-7.pdf",
       dpi = 600,
       height = 2.5,
       width = 1.5)
```

```{r }
library(dplyr)
library(tidyr)

# Summarize per sample for upper and lower bins
ratio_data_long <- ratio_data %>%
  mutate(
    BinGroup = "Bin1-3",
    Value = bin1_3
  ) %>%
  dplyr::select(Sample, Condition, BinGroup, Value) %>%
  bind_rows(
    ratio_data %>%
      mutate(
        BinGroup = "Bin5-7",
        Value = bin5_7
      ) %>%
      dplyr::select(Sample, Condition, BinGroup, Value)
  ) %>%
  mutate(BinGroup = factor(BinGroup, levels = c("Bin1-3", "Bin5-7")))

```

# Fig 7i FEZF2 panels
```{r fig.height=2.5, fig.width=2.5}
stats_ratio <- ratio_data_long %>%
  group_by(BinGroup) %>%
  summarise(
    p_val = wilcox.test(Value ~ Condition)$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    p_adj = p.adjust(p_val, method = "fdr"),   # FDR correction
    p_sign = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE ~ "ns"
    ),
    x = 1.5,  # midpoint between Control and TSC
    y.position = ratio_data_long %>%
      group_by(BinGroup) %>%
      summarise(max_val = max(Value, na.rm = TRUE)) %>%
      pull(max_val) * 1.05,
    p_label = paste0(p_sign, "\np=", sprintf("%.3f", p_adj))
   )


gg_ratio_box <- ggplot(ratio_data_long, aes(x = Condition, y = Value, fill = Condition, shape = Condition)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, width = 0.6) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
  geom_text(
    data = stats_ratio,
    aes(x = x, y = y.position, label = p_label),
    inherit.aes = FALSE,
    size = 3.5,
    lineheight = 0.8,
    vjust = -0.5
  ) +
  facet_wrap(~BinGroup, scales = "free_y") +
  scale_fill_manual(values = c("Control" = "dodgerblue2", "TSC" = "deeppink3")) +
  scale_shape_manual(values = c("Control" = 16, "TSC" = 17)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.3))) +
  labs(y = "FEZF2 positive cells", x = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    strip.text = element_text(face = "bold", size = 9),
    axis.text.x = element_blank()
  )

#Plot
gg_ratio_box

ggsave(plot = gg_ratio_box,
       filename = "figures/RNAscope_Boxplot_FEZF2_positive_cells_1-3+1-7_conditions.pdf",
       dpi = 600,
       height = 2.5,
       width = 2.5)


```
```{r}
sessionInfo()
```

