---
title: "Postmortem vs Neurosurgery Gene Signatures: Impact on DE Pathways"
output: html_notebook
---

# LIBRARIES
```{r setup, include=FALSE}
library(pagoda2)
library(conos)
library(cacoa)
library(dplyr)
library(readr)
library(purrr)
library(clusterProfiler)
library(ggplot2)
library(tidyr)
library(parallel)
library(fgsea)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(GO.db)
```

```{r message=FALSE, include=FALSE}
# Check if 'figure' folder exists, if not, create it
if (!dir.exists("figures")) {
  dir.create("figures")
  cat("Created 'figures' folder\n")
} else {
  cat("'figures' folder already exists\n")
}

```


## Load postmortem + surgery gene sets
```{r}
signatures <- readr::read_csv2("gene_signatures_neurosurgery_postmortem.csv")

surgery_genes <- unique(signatures$`Neurosurgical DEG`) %>% na.omit()
postmortem_genes <- unique(signatures$`Postmortem DEG`) %>% na.omit()
```

## Load data objects
```{r eval=FALSE, include=FALSE}
cao <- readRDS("../../5.neurons/cao_neurons.rds")
annotation <- readRDS("../../../TSC2024_data_files/5.neurons/annotation_neurons.rds")
```

## Compute DE per cell type (l2)
```{r eval=FALSE, include=FALSE}
cao$estimateDEPerCellType(cell.groups = annotation$l2, name = "de_l2")
de <- cao$test.results$de_l2

```


## With Z-scores
```{r}
rank_metric <- function(df) {
  ranks <- df$stat      # Wald Z-statistic
  names(ranks) <- df$Gene
  sort(ranks, decreasing = TRUE)
}
```



## fgsea for postmortem + surgery genesets 
```{r}
gene_sets <- list(Postmortem = postmortem_genes, Surgery = surgery_genes)

fgsea_results_list <- lapply(names(de), function(ct) {
  df <- de[[ct]]$res
  ranks <- rank_metric(df)
  
  TSC_ranks <- ranks[ranks > 0]
  Control_ranks <- ranks[ranks < 0]
  
  fgsea_TSC <- fgseaMultilevel(pathways = gene_sets, stats = TSC_ranks, scoreType = "pos") %>%
    as_tibble() %>%
    mutate(GroupUp = "TSC_up",
           OverlapGenes = map(pathway, ~paste(intersect(names(TSC_ranks), gene_sets[[.x]]), collapse = ", ")))
  
  fgsea_Control <- fgseaMultilevel(pathways = gene_sets, stats = Control_ranks, scoreType = "neg") %>%
    as_tibble() %>%
    mutate(GroupUp = "Control_up",
           OverlapGenes = map(pathway, ~paste(intersect(names(Control_ranks), gene_sets[[.x]]), collapse = ", ")))
  
  bind_rows(fgsea_TSC, fgsea_Control) %>% mutate(CellType = ct)
})

fgsea_results <- bind_rows(fgsea_results_list) %>%
  dplyr::select(CellType, GroupUp, pathway, ES, NES, pval, padj, size, OverlapGenes)
```

## Calculate GSEA per cell type (l2)
```{r eval=FALSE, include=FALSE}
org <- org.Hs.eg.db
cao$estimateOntology(de.name = "de_l2", type="GSEA", ignore.cache = T, org.db = org, n.top.genes = 500,name="GSEA_l2")
```

```{r}
saveRDS(cao, "cao_post_surg.rds",compress=F)
```

## Extract leading edge genes from cao
```{r}
extract_leadingEdge <- function(gsea_obj) {
  if (is.null(gsea_obj)) return(NULL)
  df <- as.data.frame(gsea_obj@result)
  df %>%
    dplyr::select(GOID = ID, padj = p.adjust, NES, core_enrichment) %>%
    mutate(LeadingGenes = strsplit(as.character(core_enrichment), "/"))
}
```

## Extract BP per cell typefrom gsea
```{r}
prior_pathways_long <- map_dfr(
  names(cao$test.results$GSEA_l2$res),
  function(ct) {
    ct_obj <- cao$test.results$GSEA_l2$res[[ct]]
    gsea_res <- ct_obj[["BP"]]
    if (is.null(gsea_res)) return(NULL)
    extract_leadingEdge(gsea_res) %>%
      mutate(CellType = ct, Ontology = "BP")
  }
)
```

## Remove redundant parent GO terms 
```{r}

bp_parents <- as.list(GOBPPARENTS)

parents_flat <- prior_pathways_long$GOID %>%
  map(~ bp_parents[[.x]]) %>% 
  compact() %>% 
  unlist() %>%
  unique()

# Keep only the most specific GOIDs (children)
prior_pathways_long <- prior_pathways_long %>%
  filter(!GOID %in% parents_flat)

```

## Compute overlap with top pathways
```{r}
fgsea_enriched_long <- fgsea_results %>%
  filter(!is.na(OverlapGenes) & OverlapGenes != "") %>%
  mutate(Genes = strsplit(as.character(OverlapGenes), ",\\s*")) %>%
  dplyr::select(CellType, GroupUp, FGSEAGeneSet = pathway, Genes) %>%
  tidyr::unnest(Genes)

fgsea_sig_by_set <- fgsea_enriched_long %>%
  group_by(CellType, FGSEAGeneSet, GroupUp) %>%
  summarise(SigGenes = list(unique(Genes)), .groups="drop")

combined_leading_edge <- prior_pathways_long %>%
  filter(CellType != "TBD") %>% 
  left_join(fgsea_sig_by_set, by = "CellType") %>%
  rowwise() %>%
  mutate(
    SigGenesInLeadingEdge = list(intersect(LeadingGenes, SigGenes)),
    N_SigInLeadingEdge = length(SigGenesInLeadingEdge),
    PropSigInLeadingEdge = N_SigInLeadingEdge / length(LeadingGenes)
  ) %>%
  filter(N_SigInLeadingEdge > 0) %>%
  ungroup()

signature_driven_pathways <- combined_leading_edge %>%
  mutate(NES_direction = ifelse(NES >= 0, "Up", "Down"))
```

## Look for mTOR signaling in signature genes
```{r}
library(msigdbr)
library(dplyr)

reactome_sets <- msigdbr(species = "Homo sapiens", 
                         collection = "C2", 
                         subcollection = "CP:REACTOME")

mtor_reactome <- reactome_sets %>%
  filter(grepl("MTOR", gs_name, ignore.case = TRUE))

mtor_related <- reactome_sets %>%
  filter(grepl("MTOR|AKT|PI3K|TSC|RPS6|EIF4E|AUTOPHAGY", gs_name, ignore.case = TRUE)) %>%
  dplyr::select(gs_name, gene_symbol) %>%
  distinct()

canonical_mtor <- mtor_related %>%
  filter(
    !grepl("CANCER|SARS|ERBB|NTRK|PTK6|ABERRANT|MET|CD28|CONSTITUTIVE|NEGATIVE_REGULATION|ERYTH|PI3KGAMMA", gs_name, ignore.case = TRUE)
  )

mtor_canonical_genes <- canonical_mtor %>%
  pull(gene_symbol) %>%
  unique()

length(mtor_canonical_genes)
```

```{r}
combined_leading_edge_table <- combined_leading_edge %>%
  rowwise() %>%
  mutate(
    mTOR_in_Leading = list(intersect(LeadingGenes, mtor_canonical_genes)),
    mTOR_in_SignatureLeading = list(intersect(SigGenesInLeadingEdge, mtor_canonical_genes)),
    N_mTOR_in_Leading = length(mTOR_in_Leading),
    N_mTOR_in_SignatureLeading = length(mTOR_in_SignatureLeading)
  ) %>%
  ungroup()

combined_leading_edge_table <- combined_leading_edge_table %>%
  rowwise() %>%
  mutate(
    # Genes in canonical mTOR pathway
    LeadingGenesInMTOR = list(intersect(LeadingGenes, mtor_canonical_genes)),
    N_LeadingGenesInMTOR = length(LeadingGenesInMTOR),
    
    # Signature genes in canonical mTOR
    SigGenesInMTOR = list(intersect(SigGenesInLeadingEdge, mtor_canonical_genes)),
    N_SigGenesInMTOR = length(SigGenesInMTOR)
  ) %>%
  ungroup()
```

```{r}
combined_leading_edge_table <- combined_leading_edge_table %>%
  mutate(
    Prop_LeadingInMTOR = N_LeadingGenesInMTOR / length(LeadingGenes),
    Prop_SigInMTOR = N_SigGenesInMTOR / N_SigInLeadingEdge
  )

```

## Convert GOID to descriptive terms and filter TBD
```{r}
library(dplyr)

# Map GO IDs → descriptive terms
go_mapping <- AnnotationDbi::select(
  GO.db,
  keys = unique(combined_leading_edge_table$GOID),
  columns = c("GOID", "TERM"),
  keytype = "GOID"
) %>%
  distinct(GOID, .keep_all = TRUE)

combined_leading_edge_table <- combined_leading_edge_table %>%
  left_join(go_mapping, by = "GOID")

# Filter only relevant BP pathways and TBD cell types
filtered_pathways <- combined_leading_edge_table %>%
  filter(Ontology == "BP", !CellType %in% "TBD") %>%
  mutate(
    NES_direction = ifelse(NES >= 0, "Up", "Down")
  )
```

## Select top 20 pathways
```{r}
top_pathways_keys <- filtered_pathways %>%
  group_by(GroupUp, GOID) %>%
  summarise(
    NES_mean = mean(as.numeric(NES), na.rm = TRUE),
    padj_min = min(as.numeric(padj), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # Keep only significant pathways
  filter(padj_min < 0.05) %>%
  arrange(GroupUp, padj_min, desc(abs(NES_mean))) %>%
  group_by(GroupUp) %>%
  slice_head(n = 15) %>%   # top 10 significant per GroupUp
  ungroup()

top_pathways <- filtered_pathways %>%
  semi_join(top_pathways_keys, by = c("GroupUp", "GOID"))

```



# Extended data fig. 18
```{r fig.height=12, fig.width=15}
library(patchwork)
library(ggplot2)
library(dplyr)
library(forcats)

celltypes <- c("L2_CUX2","L2_3_CUX2","L4_RORB","L5_6","ID2","PVALB","SST","VIP")

# Ensure consistent ordering of pathways
heatmap_data <- top_pathways %>%
  rowwise() %>%
  mutate(
    TERM_f = fct_reorder(TERM, PropSigInLeadingEdge, .fun = max)
  ) %>%
  ungroup()

# Consistent CellType ordering
celltype_levels <- unique(heatmap_data$CellType)
heatmap_data <- heatmap_data %>%
  mutate(CellType = factor(CellType, levels = celltype_levels))

# ---- NEW: compute % mTOR genes and text labels ----
heatmap_data <- heatmap_data %>%
  mutate(
    PercentMTOR = 100 * N_SigGenesInMTOR / N_SigInLeadingEdge,
    Label = ifelse(PropSigInLeadingEdge > 0.25,
                   paste0(round(PercentMTOR, 1), "%"),
                   NA)
  )

# Split data
postmortem_data <- heatmap_data %>% filter(FGSEAGeneSet == "Postmortem")
surgery_data <- heatmap_data %>% filter(FGSEAGeneSet == "Surgery")

# ---- Shared scales for NES dot plots ----
nes_fill <- scale_fill_distiller(
  palette = "PuOr",
  limits = range(c(-3, 3), na.rm = TRUE),
  name = "NES",
  type = "div",
  direction = -1)

padj_size <- scale_size_continuous(
  range = c(3,10),
  name = "-log10(padj)",
  limits = range(-log10(heatmap_data$padj),na.rm = TRUE))

# ---- PropSigInLeadingEdge heatmaps ----
p1_post <- ggplot(postmortem_data, aes(x = CellType, y = TERM_f, fill = PropSigInLeadingEdge)) +
  geom_tile(color = "white") +
  # ---- ADD percentage text ----
  geom_text(aes(label = Label), color = "black", size = 3.5) +
  scale_fill_distiller(palette = "OrRd", direction = 1, limits = c(0, 1),
                     name = "Proportion of\nsignature genes\nin leading genes") +
  scale_x_discrete(limits = celltypes) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), axis.title.x = element_blank(),
        axis.text.y = element_text(face="bold", colour = "darkslategrey"),
        plot.title = element_text(size=12, hjust=0.5, face="bold")) +
  labs(title = "Overlap with postmortem signature genes", x = "Cell type", y = NULL)

p1_surg <- ggplot(surgery_data, aes(x = CellType, y = TERM_f, fill = PropSigInLeadingEdge)) +
  geom_tile(color = "white") +
  # ---- ADD percentage text ----
  geom_text(aes(label = Label), color = "black", size = 3.5) +
  scale_fill_distiller(palette = "OrRd", direction = 1, limits = c(0, 1),
                     name = "Proportion of\nsignature genes\nin leading genes") +
  scale_x_discrete(limits = celltypes) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", colour = "black"),
        axis.title.x = element_blank(),
        axis.text.y = element_text(face="bold", colour = "darkslategrey"),
        plot.title = element_text(size=12, hjust=0.5, face="bold")) +
  labs(title = "Overlap with surgery signature genes", x = "Cell type", y = NULL)

# ---- NES dot plots ----
p2_post <- ggplot(postmortem_data, aes(x = CellType, y = TERM_f, fill = NES, size = -log10(padj))) +
  geom_point(shape = 21, color = "black") +
  nes_fill + padj_size +
  scale_y_discrete(position = "right") +
  scale_x_discrete(limits = celltypes) +
  theme_minimal() +
  theme(axis.text.y = element_blank(), 
        axis.title.y.right = element_text(angle = 90, vjust = 0.5),
        axis.text.x = element_blank(), axis.title.x = element_blank(),
        plot.title = element_text(size=12, hjust=0.5, face="bold")) +
  labs(title = "Pathway enrichment", x = "Cell type", y = "Postmortem")

p2_surg <- ggplot(surgery_data, aes(x = CellType, y = TERM_f, fill = NES, size = -log10(padj))) +
  geom_point(shape = 21, color = "black") +
  nes_fill + padj_size +
  scale_y_discrete(position = "right") +
  scale_x_discrete(limits = celltypes) +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.title.y.right = element_text(angle = 90, vjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", colour = "black"),
        axis.title.x = element_blank(),
        plot.title = element_text(size=12, hjust=0.5, face="bold")) +
  labs(title = "Pathway enrichment", x = "Cell type", y = "Surgery")

# ---- Combine with patchwork and collect legends ----
combined_plot <- (p1_post | p2_post) / (p1_surg | p2_surg) + 
  plot_layout(guides = "collect") & theme(legend.position = "right")

combined_plot + 
  plot_annotation(
    caption = "Cell type",
    theme = theme(
      plot.caption = element_text(size = 15, hjust = 0.58, vjust = 3)
    )
  ) 

filename <- "figures/Postmortem_surgery_overlapsiginleading_percent+nes_vs_sign.pdf"
pdf.width = 12
pdf.height = 12
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

```{r}
sessionInfo()
```

