---
title: "snRNA-seq TSC2: Case-control analysis - Neurons"
output: html_notebook
author: "Frederik Nørby Friis Sørensen, University of Copenhagen"
date: "April 2026"
---

## LIBS
```{r setup}
library(dplyr)
library(magrittr)
library(conos)
library(pagoda2)
library(scHelper)
library(cacoa)
library(ggplot2)
library(ggrastr)
library(cowplot)
library(HDF5Array)
library(Seurat)
set.seed(1)
```

```{r message=FALSE, include=FALSE}
# Check if 'data' folder exists, if not, create it
if (!dir.exists("data")) {
  dir.create("data")
  cat("Created 'data' folder\n")
} else {
  cat("'data' folder already exists\n")
}

# Check if 'figure' folder exists, if not, create it
if (!dir.exists("figures")) {
  dir.create("figures")
  cat("Created 'figures' folder\n")
} else {
  cat("'figures' folder already exists\n")
}

```


# Load basic objects
```{r}
cms<- readRDS("../../TSC2024_data_files/3.all_cells/cms_final_filter.rds")
annotation <- readRDS("../../TSC2024_data_files/3.all_cells/annotation_all_2024.rds")
```

## Subset neurons
```{r}
cms %<>% lapply(function(x) x[,colnames(x) %in% names(annotation$l1[annotation$l1 %in% c("Inhibitory neurons","Excitatory neurons")])])
```

```{r}
saveRDS(cms, "cms_neurons.rds", compress=F)
```

```{r}
cms <- readRDS("cms_neurons.rds")
```

```{r eval=FALSE, include=FALSE}
con_neurons<- quickConos(cms,
                  names(cms), 
                  make.geneknn = F,
                  n.odgenes = 3e3,
                  nPcs = 100,
                  min.cells.per.gene = 0, 
                  min.transcripts.per.cell = 0, 
                  ncomps = 40,
                  get.tsne = F,
                  n.cores.p2=100,
                  n.cores.con=100)
con_neurons <- con_neurons$con

saveRDS(object = con_neurons, file = "con_neurons.rds", compress = F)
```

```{r}
con_neurons <- readRDS("con_neurons.rds")
```
# Isolate inhibitory cells to make annotation with higher granularity

```{r}
cms_inh <- cms
cms_inh %<>% lapply(function(x) x[,colnames(x) %in% names(annotation$l1[annotation$l1 %in% c("Inhibitory neurons")])])

```

```{r}
saveRDS(cms_inh,"cms_inhibitory.rds", compress=F)
```

## Make conos object for inhib cells

```{r eval=FALSE, include=FALSE}
con_inh<- quickConos(cms_inh,
                  names(cms_inh), 
                  make.geneknn = F,
                  n.odgenes = 3e3,
                  nPcs = 100,
                  min.cells.per.gene = 3, 
                  min.transcripts.per.cell = 200, 
                  ncomps = 40,
                  get.tsne = T,
                  n.cores.p2=100,
                  n.cores.con=100)
con_inh <- con_inh$con

saveRDS(object = con_inh, file = "con_inhibitory.rds", compress = F)
```

# Isolate excitatory cells 

```{r}
cms_exci <- cms
cms_exci %<>% lapply(function(x) x[,colnames(x) %in% names(annotation$l1[annotation$l1 %in% c("Excitatory neurons")])])
```

```{r}
saveRDS(cms_exci,"cms_excitatory.rds", compress=F)
```

## Make conos object for excitatory neurons

```{r eval=FALSE, include=FALSE}
con_exci<- quickConos(cms_exci,
                  names(cms_exci), 
                  make.geneknn = F,
                  n.odgenes = 3e3,
                  nPcs = 100,
                  min.cells.per.gene = 3, 
                  min.transcripts.per.cell = 200, 
                  ncomps = 40,
                  get.tsne = T,
                  n.cores.p2=100,
                  n.cores.con=100)
con_exci <- con_exci$con

saveRDS(object = con_exci, file = "con_excitatory.rds", compress = F)
```

# Load neuron annotation
```{r}
annotation_n <- readRDS("../../TSC2024_data_files/5.neurons/annotation_neurons.rds")
```

```{r}
dir.create("figures/cacoa")
```

# Cacoa Control vs lesion - neurons
```{r eval=FALSE, include=FALSE}
##Cacoa Control vs lesion
cao <- Cacoa$new(data.object = con_neurons, sample.groups = c(rep("Control",6), rep("TSC2",11)) %>% setNames(names(con_neurons$samples)), cell.groups = annotation_n$l4, ref.level = "Control", target.level = "TSC2", n.cores = 100)
```


```{r eval=FALSE, include=FALSE}
saveRDS(cao, "cao_neurons.rds", compress=F)
```

```{r}
bs <- if (length(levels(cao$cell.groups)) > 20) 10 else 16
cao$plot.theme <- theme_bw(base_size = bs) + 
  theme(plot.title=element_text(hjust = 0.5), 
        legend.background=element_rect(fill=alpha("white", 0.2)),
        legend.text=element_text(size=12), 
        legend.margin=margin(6, 6, 4, 1, 'pt'),
        plot.margin=margin())


cao$cell.groups.palette <- levels(cao$cell.groups) %>% 
  {setNames(sample(brewerPalette("Paired")(length(.))), .)}

cao$sample.groups.palette <- c("#d73027", "#4575b4") %>% 
  setNames(c(cao$target.level, cao$ref.level))
```

```{r}
# sorted cell type names, with the largest cell types first

example.celltypes <- names(sort(table(cao$cell.groups),decreasing=T))
top1 <- example.celltypes[c(1,4)]
top2 <- example.celltypes[c(1,5,6,9,13,17,31,34)]
 
cells.to.contour <- top2
cell.to.go <- example.celltypes[2]

alpha <- 0.2; size <- 0.1;
```

## Estimate shifts and loadings
```{r}
cao$estimateExpressionShiftMagnitudes()
cao$estimateCellLoadings()
```

## Expression distance
### Fig. 3a
```{r}
p3 <- cao$plotExpressionShiftMagnitudes(type = "box",show.pvalues = "adjusted")
p3
 
 iqr = function(z, lower = 0.25, upper = 0.75) {
  data.frame(
    y = median(z),
    ymin = quantile(z, lower),
    ymax = quantile(z, upper)
  )
}


p_ed <- p3$data %>% filter(Type !="TBD") %>% ggplot( aes( y = value, x = Type, color = Type)) +
  geom_violin(scale = "width", 
                #fill = "#ededed", 
                color = NA, aes(fill = Type), alpha = 0.2, 
              trim = F) +
  theme_light() + theme(legend.position = "none", axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), panel.grid.minor = element_blank(), panel.grid.major = element_blank()) + scale_color_manual(values = cao$cell.groups.palette)+ scale_fill_manual(values = cao$cell.groups.palette) +
  geom_hline(yintercept = 0, linewidth = 0.1) +
  stat_summary(fun.data = iqr, geom = "errorbar", show.legend = F, aes(color = Type), shape = 23, stroke = 0.5, alpha = 0.9,color = "grey30", width = 0.25) + 
  stat_summary(fun.data = iqr, geom = "point", show.legend = F,aes(fill = Type), shape = 23, stroke = 0.5, color = "grey30", alpha = 0.9)  + ylab(("Normalized expression distance")) + xlab("")
p_ed
```


```{r}
pdf(file = "figures/cacoa/expression_distance_new.pdf", height = 5, width = 12)
p_ed
dev.off()
```

## Cluster based compositional 

```{r}
p2 <- cao$plotCellLoadings(show.pvals = F)
p2 <- p2$data
```
### Fig. 2b 
```{r}
library(data.table)
library(ggnewscale)
p2 <- data.table(p2)

palcomp <- setNames(ifelse(p2[, ((values)), by = ind][order(-V1),V1] > 0, "deeppink3", "dodgerblue2"), p2[, ((values)), by = ind][order(-V1),ind])
palcomp2 <- setNames(ifelse(p2[, ((values)), by = ind][order(-V1),V1] > 0, "darkblue", "grey50"), p2[, ((values)), by = ind][order(-V1),ind])


p_cl <- p2 %>% filter(ind !="TBD") %>% ggplot( aes( x = values, y = ind, color = ind)) +
  geom_violin(scale = "width", 
                #fill = "#ededed", 
                color = NA, aes(fill = ind), alpha = 0.2) +
  theme_light() + theme(legend.position = "none") + scale_fill_manual(values = palcomp) +
  geom_vline(xintercept = 0, linewidth = 0.2) +
  new_scale_fill() + 
  stat_summary(fun.data = iqr, geom = "errorbar", show.legend = F, aes(color = ind), stroke = 0.6, alpha = 0.9,color = "grey30", width = 0.25) + 
  stat_summary(fun.data = iqr, geom = "point", show.legend = F,aes(fill = ind), shape = 23, stroke = 1, color = "grey30", alpha = 0.9, size = 2) +  scale_fill_manual(values = palcomp) + xlab("Separation coefficient") + 
  ylab("") +
  xlim(-1.0,1.0)
  
p_cl
```
```{r}
pdf(file = "figures/cacoa/Cell_loadings_new.pdf", height = 12, width = 7)
p_cl
dev.off()
```

#Cluster-free compositional
##Estimate cell density
```{r}
cao$estimateCellDensity(method='graph', verbose=F)
cao$estimateDiffCellDensity(type='wilcox',verbose=F)
```

```{r}
p1 <- cao$plotEmbedding(color.by='cell.groups', alpha=0.3, size=0.05, 
                    plot.na=FALSE, show.legend=FALSE, font.size=4) %>% rasterise(dpi=900)+ theme(panel.grid.major = element_blank(), 
               panel.grid.minor = element_blank(), panel.border = element_blank())

    
p2 <- cao$plotDiffCellDensity(legend.position=c(0, 1)) %>% rasterize(dpi=900)+ theme(panel.grid.major = element_blank(), 
               panel.grid.minor = element_blank(), panel.border = element_blank())
```


### Fig. 1h
```{r fig.height=8, fig.width=8, warning=FALSE}
pdf(file = "figures/cacoa/Annotation.pdf", height = 10, width = 10)
p1 + theme(plot.margin=unit(c(0.4,0.4,0.4,0.4),"cm"))
dev.off()
```
### Fig. 2d

```{r fig.height=8, fig.width=16, warning=FALSE}
pdf(file = "figures/cacoa/diff_cell_density_clear.pdf", height = 8, width = 16)
p1+p2 + theme(plot.margin=unit(c(0.4,0.4,0.4,0.4),"cm"))
dev.off()
```


#Expression changes, cluster-free
```{r}
cao$estimateClusterFreeExpressionShifts(
  gene.selection="expression", min.n.between=3, min.n.within=3
)
```

### Fig. 3c
```{r}
p3 <- cao$plotClusterFreeExpressionShifts(cell.groups = annotation$l4,
                                          legend.position=c(0, 1),
                                          plot.na =F,
                                          font.size=4,
                                          size=0.2,
                                          alpha=0.6,
                                          build.panel = F,)[[2]] %>% rasterize(dpi=900) + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank())
```

```{r fig.height=6, fig.width=6}
pdf(file = "figures/cacoa/Cluster_free_expression.pdf", height = 10, width = 10)
p3 %>% rasterize(dpi=900)
dev.off()
```

```{r}
saveRDS(cao, "cao_neurons.rds", compress=F)
```
```{r}
cao <- readRDS("cao_neurons.rds")
```

#DE and GO
```{r}
cao$estimateDEPerCellType(
  independent.filtering=T, test='DESeq2.Wald', verbose=T ,n.cores = 200,seed.resampling = 42,min.cell.count = 2)
```

### Make DE table for all neurons with significance threshold of alpha = 0.05

```{r}
library(data.table)
de_neurons <- rbindlist(lapply(cao$test.results$de, function(x) x$res), idcol = "CellType")
de_neurons <- de_neurons[padj <= 0.05]
de_neurons
write.csv2(de_neurons,file = "differential_expression_neurons.csv")
```

## Volcanoplots
### Extended data fig. 10
```{r fig.height=21.15, fig.width=15}
# Excitatory/principal neurons
de <- cao$test.results$de
exci <- c(
  "L2_CUX2_LAMP5_PDGFD","L2_3_CUX2_FREM3_NTNG1", "L3_CUX2_SPOCK3_PRSS12", 
  "L4_RORB_SCHLAP1_ARHGAP15", "L4_RORB_SCHLAP1_MET", "L4_RORB_SCHLAP1_MME", 
  "L5_6_FEZF2_LRRK1_PCP4", "L5_6_FEZF2_LRRK1_SEMA3E", "L5_6_FEZF2_TLE4_ABO", 
  "L5_6_FEZF2_TLE4_HTR2C", "L5_6_FEZF2_TLE4_SCUBE1", "L5_6_THEMIS_NR4A2_NTNG2", 
  "L5_6_THEMIS_SEMA3A"
)

volc_exci <- cao$plotVolcano(
  xlim = c(-3, 3),
  ylim = c(0, 5),
  lf.cutoff = 1,
  cell.types = exci,
  n.col = 3,
  color.var = "Za",
  pointSize = 0.5,
  colAlpha = 0.25,
  raster = TRUE
) + 
  theme(plot.margin = unit(c(1, 1, 1, 1), "cm"))

volc_exci

ggsave(
  filename = "figures/cacoa/Volcano_plot_excitatory_neurons.png",
  plot = volc_exci,
  width = 15,
  height = 21.15,
  dpi = 300,
  bg = "white"
)
```

### Extended data fig. 11
```{r fig.height=28, fig.width=20}
# Inhibitory neurons
inh <- c(  "ID2_CHRNA7_CNR1", "ID2_CHRNA7_NCKAP5", "ID2_CHRNA7_PAX6", "ID2_LAMP5_CRH", "ID2_LAMP5_MEIS2", "ID2_LAMP5_NMBR", "ID2_LAMP5_NOS1",
           "VIP_ABI3BP", "VIP_CBLN1", "VIP_CRH", "VIP_NRG1", "VIP_SEMA3C", "VIP_TRPM3_SSTR1", "VIP_TYR",
           "PVALB_BTBD11", "PVALB_LGR5", "PVALB_MEPE", "PVALB_UNC5B",
           "SST_CALB1", "SST_ISOC_THSD7B", "SST_NOS1", "SST_NPY", "SST_STK32A", "SST_TAC3", "SST_TH")

volc_inh <- cao$plotVolcano(xlim=c(-3, 3), ylim=c(0, 5), lf.cutoff=1, cell.types = inh,n.col = 4, color.var = "Za", pointSize=0.5,colAlpha = 0.25, raster=T)+ theme(plot.margin=unit(c(1,1,1,1),"cm"))
volc_inh
ggsave(filename = "figures/cacoa/Volcano_plot_inhbitory_neurons.png",plot = volc_inh,width = 20,height = 28,dpi = 300,bg="white")
```

## Estimate GSEA

```{r}
library(org.Hs.eg.db)
org <- org.Hs.eg.db

cao$estimateOntology(type="GSEA",
                     org.db = org,
                     n.top.genes = 500,
                     name="GSEA")
```

```{r eval=FALSE, include=FALSE}
saveRDS(cao,"cao_neurons.rds", compress=F)
cao <- readRDS("cao_neurons.rds")
```

## GSEA heatmap down
```{r fig.height=6.5, fig.width=11}
cao$plotOntologyHeatmap(name="GSEA", genes="down",top.n = 25, legend.title="-log10(p-value)", min.genes=25,p.adj = 0.05,max.log.p = 10,)+ theme(plot.margin=unit(c(0.5,0.5,0.5,0.5),"cm"))
filename = "figures/cacoa/GSEA_down_HM.pdf"
pdf.width = 8.5
pdf.height = 4.5
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)

```

##GSEA heatmap up 
```{r fig.height=6.5, fig.width=9}
cao$plotOntologyHeatmap(name="GSEA", genes="up",top.n = 25, legend.title="-log10(p-value)", min.genes=25,p.adj = 0.05,max.log.p = 15)+ theme(plot.margin=unit(c(0.5,0.5,0.5,0.5),"cm"))
filename = "figures/cacoa/GSEA_UP_HM.pdf"
pdf.width = 10
pdf.height = 4.5
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```
# GSEA collapsed heatmap/dotplots
## GSEA UP
### Fig. 4a
```{r fig.height=8, fig.width=10}
cao$plotOntologyHeatmapCollapsed(name="GSEA", genes="up", n=30, clust.method="ward.D", min.genes=20,p.adj = 0.05)+ theme(plot.margin=unit(c(0.5,0.5,0.5,0.5),"cm"))
filename = "figures/cacoa/GSEA_UP_HM_collapsed.pdf"
pdf.width = 10
pdf.height = 8
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

## GSEA DOWN
### Fig. 4b
```{r fig.height=8, fig.width=10}
cao$plotOntologyHeatmapCollapsed(name="GSEA", genes="down", n=30, clust.method="ward.D", min.genes=20, p.adj = 0.05)+ theme(plot.margin=unit(c(0.5,0.5,0.5,0.5),"cm"))
filename = "figures/cacoa/GSEA_DOWN_HM_collapsed.pdf"
pdf.width = 10
pdf.height = 8
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

## GSEA glycolysis vs fatty acids
### Fig. 4c
```{r fig.height=5, fig.width=10}
cao$plotOntologyHeatmapCollapsed(name="GSEA",description.regex = "fatt|trig|glycoly|aero|lipid|lipop|glycoge|pentose|fructose|galactose",max.log.p = 5,clust.method="ward.D",p.adj = 0.05, genes="all")+theme(plot.margin=unit(c(0.5,0.5,0.5,0.5),"cm"))
filename = "figures/cacoa/GSEA_glycolysis_fattyacids.pdf"
pdf.width = 10
pdf.height = 6
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

## GSEA Axon/synapse development
### Fig. 4d
```{r fig.height=5, fig.width=10}
cao$plotOntologyHeatmapCollapsed(name="GSEA",description.regex = "synap",max.log.p = 5,n = 10,clust.method="ward.D",p.adj = 0.05, genes="all")+theme(plot.margin=unit(c(0.5,0.5,0.5,0.5),"cm"))

filename = "figures/cacoa/GSEA_dotplot_synap_develop.pdf"
pdf.width = 10
pdf.height = 6
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

##Pull out GO/GSEA test results
```{r}
library(data.table)
rbindlist(lapply(cao$test.results$GSEA$res, function(x) x$BP@result), idcol = "CellType")
rbindlist(lapply(cao$test.results$GSEA$res, function(x) x$CC@result), idcol = "CellType") 
rbindlist(lapply(cao$test.results$GSEA$res, function(x) x$MF@result), idcol = "CellType")
```

```{r}
ontologies<- rbindlist(list(rbindlist(lapply(cao$test.results$GSEA$res, function(x) x$BP@result), idcol = "CellType"),
                rbindlist(lapply(cao$test.results$GSEA$res, function(x) x$CC@result), idcol = "CellType"), 
                rbindlist(lapply(cao$test.results$GSEA$res, function(x) x$MF@result), idcol = "CellType")), idcol = "TypeOfOntology")
```
```{r}
axon_development <- ontologies %>% filter(grepl('^axon development', Description))
```

```{r}
mTOR <- ontologies %>% filter(grepl('^TOR', Description))
mTOR
write.csv2(mTOR, "mTOR.csv")
```

```{r}
neuron_migration <- ontologies %>% filter(grepl('neuron migrat|neuronal migrat', Description))
neuron_migration
write.csv2(neuron_migration, "neuron_migration.csv")
```

```{r}
neuron_develop <- ontologies %>% filter(grepl('neuro|develop', Description))
neuron_develop
write.csv2(neuron_develop, "neuron_develop.csv")
```
```{r}
neuron_synap <- ontologies %>% filter(grepl('synap', Description))
neuron_synap
write.csv2(neuron_synap, "neuron_synap.csv")
```

```{r}
neuron_matur <- ontologies %>% filter(grepl('neuron matur', Description))
neuron_matur
write.csv2(neuron_matur, "neuron_matur.csv")
```

#fatt|trig|glycoly|aero|lipid|lipop|glycoge|pentose|fructose|galactose
```{r}
fatty_acid <- ontologies %>% filter(grepl('lipid metabolic|long-chain-fatty|lipid catabolic|fatty-acyl|lipoprotein|fatty acid', Description))
fatty_acid
write.csv2(fatty_acid, "fatty_acid.csv")
```
```{r}
glycolytic <- ontologies %>% filter(grepl('glycolysis|pentose|fructose|glycogen|anaerobic gly|aerobic gly|glycolytic|pyruvate', Description))
glycolytic
write.csv2(glycolytic, "glycolytic.csv")
```

```{r}
electron <- ontologies %>% filter(grepl('electron', Description))
electron
write.csv2(electron, "electron.csv")
```


# Glutamate receptor pathway dotplot
```{r}
de <- cao$test.results$de
```

### Extended data fig. 13
```{r fig.height=12, fig.width=15}
# Define the markers
markers <- c("RELN", "GRIA1", "GRIK1", "GRIK2", "GRID2", "SHISA6", "GRIA3", "GRM7", "GRID1",
             "RASGRF1", "GRIN2A", "NLGN1", "PTK2B", "OPRM1", "CACNG2", "HOMER1", "PLCB1", 
             "GNAQ", "GRM5", "GRIA2", "NETO1", "DAPK1", "GRIA4", "NETO2", "GRM1", "GRM8", 
             "GRIK3", "GRIK4", "MINK1", "SHANK1", "SHISA9", "GRIN2B", "GRIK5", "CACNG3", 
             "PPARGC1A", "CPEB4", "GRIN3A", "TIAM1", "GRIN2D", "CNIH3", "GSG1L", "SHANK2", 
             "CDK5R1", "PRNP", "APP") %>% rev()

# Define the cell types in the desired order
cell_types <- c("L2_CUX2_LAMP5_PDGFD","L2_3_CUX2_FREM3_NTNG1", "L3_CUX2_SPOCK3_PRSS12", 
                "L4_RORB_SCHLAP1_ARHGAP15", "L4_RORB_SCHLAP1_MET", "L4_RORB_SCHLAP1_MME", 
                "L5_6_FEZF2_LRRK1_PCP4", "L5_6_FEZF2_LRRK1_SEMA3E", "L5_6_FEZF2_TLE4_ABO", 
                "L5_6_FEZF2_TLE4_HTR2C", "L5_6_FEZF2_TLE4_SCUBE1", "L5_6_THEMIS_NR4A2_NTNG2", 
                "L5_6_THEMIS_SEMA3A", "ID2_CHRNA7_CNR1", "ID2_CHRNA7_NCKAP5", "ID2_CHRNA7_PAX6",
                "ID2_LAMP5_CRH", "ID2_LAMP5_MEIS2", "ID2_LAMP5_NMBR", "ID2_LAMP5_NOS1", 
                "VIP_ABI3BP", "VIP_CBLN1", "VIP_CRH", "VIP_NRG1", "VIP_SEMA3C", "VIP_TRPM3_SSTR1", "VIP_TYR",
                "PVALB_BTBD11", "PVALB_LGR5", "PVALB_MEPE", "PVALB_UNC5B",
                "SST_CALB1", "SST_ISOC_THSD7B", "SST_NOS1", "SST_NPY", "SST_STK32A", "SST_TAC3", "SST_TH"
                )

# Combine all data frames in the list `de` into one data frame
combined_df <- bind_rows(lapply(de, function(x) {
  df <- x$res
  df$gene <- rownames(df)
  df %>% filter(gene %in% markers)
}), .id = "celltype")

combined_df <- combined_df %>%
  filter(!celltype %in% c("Neu_dev_THY1", "TBD"))

# Convert gene column to factor with levels in the specified order
combined_df$gene <- factor(combined_df$gene, levels = markers)

# Convert celltype column to factor with levels in the specified order
combined_df$celltype <- factor(combined_df$celltype, levels = cell_types)


# Create the dot plot
ggplot(combined_df, aes(x = gene, y = celltype, size = CellFrac, color = Za)) +
  geom_point() +
  theme_classic() +
  coord_flip() +
  scale_color_distiller(palette = "PuOr",limits = c(-6, 6), oob = scales::squish) +
  labs(title = "Differential Expression of Glutamatergic Receptor Signaling Genes",
       x = element_blank(),
       y = element_blank(),
       size = "Percent expressed",
       color = "Z-adjusted") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 10))

# Save the plot as a PDF
filename <- "figures/cacoa/Glutamate_receptor_complex_dotplot_z_scores.pdf"
pdf.width = 15
pdf.height = 12
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

### Fig. 3e
```{r fig.height=7, fig.width=12}
# Define the markers
markers <- c("SACM1L","CACNG2","VWC2","SHISA6","GRIA3","OLFM3","GRIA4","GRIA2","SHISA9","GRIA1","CACNG3","CNIH3","ABHD6","CTP1C","OLFM2") %>% rev()

# Define the cell types in the desired order
cell_types <- c("L2_CUX2_LAMP5_PDGFD","L2_3_CUX2_FREM3_NTNG1", "L3_CUX2_SPOCK3_PRSS12", 
                "L4_RORB_SCHLAP1_ARHGAP15", "L4_RORB_SCHLAP1_MET", "L4_RORB_SCHLAP1_MME", 
                "L5_6_FEZF2_LRRK1_PCP4", "L5_6_FEZF2_LRRK1_SEMA3E", "L5_6_FEZF2_TLE4_ABO", 
                "L5_6_FEZF2_TLE4_HTR2C", "L5_6_FEZF2_TLE4_SCUBE1", "L5_6_THEMIS_NR4A2_NTNG2", 
                "L5_6_THEMIS_SEMA3A", "ID2_CHRNA7_CNR1", "ID2_CHRNA7_NCKAP5", "ID2_CHRNA7_PAX6",
                "ID2_LAMP5_CRH", "ID2_LAMP5_MEIS2", "ID2_LAMP5_NMBR", "ID2_LAMP5_NOS1", 
                "VIP_ABI3BP", "VIP_CBLN1", "VIP_CRH", "VIP_NRG1", "VIP_SEMA3C", "VIP_TRPM3_SSTR1", "VIP_TYR",
                "PVALB_BTBD11", "PVALB_LGR5", "PVALB_MEPE", "PVALB_UNC5B",
                "SST_CALB1", "SST_ISOC_THSD7B", "SST_NOS1", "SST_NPY", "SST_STK32A", "SST_TAC3", "SST_TH"
                )

# Combine all data frames in the list `de` into one data frame
combined_df <- bind_rows(lapply(de, function(x) {
  df <- x$res
  df$gene <- rownames(df)
  df %>% filter(gene %in% markers)
}), .id = "celltype")

combined_df <- combined_df %>%
  filter(!celltype %in% c("Neu_dev_THY1", "TBD"))

# Convert gene column to factor with levels in the specified order
combined_df$gene <- factor(combined_df$gene, levels = markers)

# Convert celltype column to factor with levels in the specified order
combined_df$celltype <- factor(combined_df$celltype, levels = cell_types)



# Create the dot plot
ggplot(combined_df, aes(x = gene, y = celltype, size = CellFrac, color = Za)) +
  geom_point() +
  theme_classic() +
  coord_flip() +
  scale_color_distiller(palette = "PuOr",limits = c(-4, 4), oob = scales::squish) +
  labs(title = "Differential Expression of AMPA Glutamatergic Receptor Signaling Genes",
       x = element_blank(),
       y = element_blank(),
       size = "Percent expressed",
       color = "Z-adjusted") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 10))

# Save the plot as a PDF
filename <- "figures/cacoa/AMPA_Glutamate_receptor_complex_dotplot_z_scores.pdf"
pdf.width = 12
pdf.height = 7
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```


#MTOR genes heatmap/dotplot
### Fig. 5a
```{r fig.height=5, fig.width=12}
# Define the markers
markers <- c("RAC1", "TSC1", "TSC2", "RHEB", "RPTOR", "RICTOR", "RRAGA", "PRR5", "MTOR", "EEF2", "BNIP3", "PXN", "ULK1", "SGK1")

# Define the cell types in the desired order
cell_types <- c("L2_CUX2_LAMP5_PDGFD","L2_3_CUX2_FREM3_NTNG1", "L3_CUX2_SPOCK3_PRSS12", 
                "L4_RORB_SCHLAP1_ARHGAP15", "L4_RORB_SCHLAP1_MET", "L4_RORB_SCHLAP1_MME", 
                "L5_6_FEZF2_LRRK1_PCP4", "L5_6_FEZF2_LRRK1_SEMA3E", "L5_6_FEZF2_TLE4_ABO", 
                "L5_6_FEZF2_TLE4_HTR2C", "L5_6_FEZF2_TLE4_SCUBE1", "L5_6_THEMIS_NR4A2_NTNG2", 
                "L5_6_THEMIS_SEMA3A", "ID2_CHRNA7_CNR1", "ID2_CHRNA7_NCKAP5", "ID2_CHRNA7_PAX6",
                "ID2_LAMP5_CRH", "ID2_LAMP5_MEIS2", "ID2_LAMP5_NMBR", "ID2_LAMP5_NOS1", 
                "VIP_ABI3BP", "VIP_CBLN1", "VIP_CRH", "VIP_NRG1", "VIP_SEMA3C", "VIP_TRPM3_SSTR1", "VIP_TYR",
                "PVALB_BTBD11", "PVALB_LGR5", "PVALB_MEPE", "PVALB_UNC5B",
                "SST_CALB1", "SST_ISOC_THSD7B", "SST_NOS1", "SST_NPY", "SST_STK32A", "SST_TAC3", "SST_TH"
                )

# Combine all data frames in the list `de` into one data frame
combined_df <- bind_rows(lapply(de, function(x) {
  df <- x$res
  df$gene <- rownames(df)
  df %>% filter(gene %in% markers)
}), .id = "celltype")

combined_df <- combined_df %>%
  filter(!celltype %in% c("Neu_dev_THY1", "TBD"))

# Ensure celltype column is a factor with the specified order
combined_df$celltype <- factor(combined_df$celltype, levels = cell_types)

# Ensure gene column is a factor with the specified order
mTOR_ordered_genes <- c("RAC1", "TSC1", "TSC2", "RHEB", "RPTOR", "RICTOR", "RRAGA", "PRR5", "MTOR", "EEF2", "BNIP3", "PXN", "ULK1", "SGK1") %>% rev()

combined_df$gene <- factor(combined_df$gene, levels = mTOR_ordered_genes)

# Convert p-values to asterisks
combined_df <- combined_df %>%
  mutate(significance = case_when(
    padj < 0.001 ~ "***",
    padj < 0.01 ~ "**",
    padj < 0.05 ~ "*",
    TRUE ~ ""
  ))

# Create the heatmap
ggplot(combined_df, aes(x = celltype, y = gene, fill = log2FoldChange)) +
  geom_tile(color = "white") +
  geom_text(aes(label = significance), color = "black", size = 3) +
  scale_fill_distiller(palette = "PuOr", limits = c(-3, 3), oob = scales::squish) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.text.y = element_text(size = 8)) +
  labs(title = "mTOR genes Differential expression",
       x = element_blank(),
       y = element_blank(),
       fill = "Log2 Fold Change") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 10))

# Save the plot as a PDF
filename <- "figures/cacoa/mTOR_genes_DE_heatmap_selected_genes.pdf"
pdf.width = 10
pdf.height = 5
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

### Extended data fig. 19
```{r fig.height=14, fig.width=12}
# Define the markers
markers <- c("RHOA","FBXW11","TSC2","TSC1","BRAF","ATG13","PML","RRAGB","RRAGA","RRAGD","RRAGC","AKT1S1","CDK2","PDCD4","ULK2","CYCS","ULK1","RAF1","SGK1","YWHAG","EIF4B","RRN3","YWHAE","YWHAB","MAPKAP1","PLD2","PLD1","YY1","NRAS","YWHAQ","AKT1","EIF4E","PPARGC1A","YWHAH","SREBF1","MAP2K2","MAP2K1","PDPK1","PRKCA","EEF2","YWHAZ","MTOR","CLIP1","RHEB","RPS6KB1","CCNE1","DDIT4","SSPO","PRR5","EIF4A1","IRS1","POLDIP3","PXN","EIF4EBP1","RB1CC1","IKBKB","RPTOR","RPS6KA1","MLST8","MAPK3","BNIP3","RAC1","HRAS","MAPK1","RICTOR","SFN","DEPTOR")

# Define the cell types in the desired order
cell_types <- c("L2_CUX2_LAMP5_PDGFD","L2_3_CUX2_FREM3_NTNG1", "L3_CUX2_SPOCK3_PRSS12", 
                "L4_RORB_SCHLAP1_ARHGAP15", "L4_RORB_SCHLAP1_MET", "L4_RORB_SCHLAP1_MME", 
                "L5_6_FEZF2_LRRK1_PCP4", "L5_6_FEZF2_LRRK1_SEMA3E", "L5_6_FEZF2_TLE4_ABO", 
                "L5_6_FEZF2_TLE4_HTR2C", "L5_6_FEZF2_TLE4_SCUBE1", "L5_6_THEMIS_NR4A2_NTNG2", 
                "L5_6_THEMIS_SEMA3A", "ID2_CHRNA7_CNR1", "ID2_CHRNA7_NCKAP5", "ID2_CHRNA7_PAX6",
                "ID2_LAMP5_CRH", "ID2_LAMP5_MEIS2", "ID2_LAMP5_NMBR", "ID2_LAMP5_NOS1", 
                "VIP_ABI3BP", "VIP_CBLN1", "VIP_CRH", "VIP_NRG1", "VIP_SEMA3C", "VIP_TRPM3_SSTR1", "VIP_TYR",
                "PVALB_BTBD11", "PVALB_LGR5", "PVALB_MEPE", "PVALB_UNC5B",
                "SST_CALB1", "SST_ISOC_THSD7B", "SST_NOS1", "SST_NPY", "SST_STK32A", "SST_TAC3", "SST_TH"
                )

# Combine all data frames in the list `de` into one data frame
combined_df <- bind_rows(lapply(de, function(x) {
  df <- x$res
  df$gene <- rownames(df)
  df %>% filter(gene %in% markers)
}), .id = "celltype")

combined_df <- combined_df %>%
  filter(!celltype %in% c("Neu_dev_THY1", "TBD"))

# Ensure celltype column is a factor with the specified order
combined_df$celltype <- factor(combined_df$celltype, levels = cell_types)

# Ensure gene column is a factor with the specified order
mTOR_ordered_genes <- c("NRAS", "HRAS", "RHOA", "RAC1", "BRAF", "RAF1", "MAP2K1", "MAP2K2", 
                        "MAPK1", "MAPK3", "AKT1", "AKT1S1", "PDPK1", "PML", "IRS1", 
                        "TSC1", "TSC2", "RHEB", "RRAGA", "RRAGB", "RRAGC", "RRAGD", 
                        "RPTOR", "RICTOR", "DEPTOR", "MLST8", "PRR5", "MTOR", 
                        "RPS6KB1", "EIF4EBP1", "EIF4A1", "EIF4E", "EIF4B", "EEF2", 
                        "RPS6KA1", "DDIT4", "SREBF1", "PDCD4", "CYCS", "BNIP3", 
                        "RB1CC1", "PXN", "FBXW11", "YY1", "PRKCA", "PLD1", "PLD2", 
                        "PPARGC1A", "ULK1", "ULK2", "ATG13", "YWHAG", "YWHAE", 
                        "YWHAQ", "YWHAB", "YWHAH", "YWHAZ", "SGK1", "MAPKAP1", 
                        "CLIP1", "SFN", "CCNE1", "RRN3", "IKBKB", "POLDIP3", "SSPO") %>% rev()

combined_df$gene <- factor(combined_df$gene, levels = mTOR_ordered_genes)

# Convert p-values to asterisks
combined_df <- combined_df %>%
  mutate(significance = case_when(
    padj < 0.001 ~ "***",
    padj < 0.01 ~ "**",
    padj < 0.05 ~ "*",
    TRUE ~ ""
  ))

# Create the heatmap
ggplot(combined_df, aes(x = celltype, y = gene, fill = log2FoldChange)) +
  geom_tile(color = "white") +
  geom_text(aes(label = significance), color = "black", size = 3) +
  scale_fill_distiller(palette = "PuOr", limits = c(-3, 3), oob = scales::squish) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.text.y = element_text(size = 8)) +
  labs(title = "mTOR genes Differential expression",
       x = element_blank(),
       y = element_blank(),
       fill = "Log2 Fold Change") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 10))

# Save the plot as a PDF
filename <- "figures/cacoa/mTOR_genes_DE_heatmap.pdf"
pdf.width = 10
pdf.height = 14
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

# Volcano Plot for GSEA axon development genes
```{r}
#Genes pulled from axon_development$core_enrichment from GSEA on ^axon development
axon_development_genes<-c("B4GALT5","BCL2","USP33","ADNP","TNR","TAOK2","CAMSAP2","TCTN1","RND2","NTRK2","UNC5D","MAP1A","SVBP","APLP2","EPHA4","CNTN1","SEMA6C","CHRNB2","UCHL1","MCF2","NIN","NRCAM","NR2E1","MAP1B","FBXO45","CDH11","TRPC5","UNC5C","EPHA5","GRM7","ZPR1","KIF5A","L1CAM","FOXG1","SLITRK2","ACTB","RAPH1","EXT1","IFRD1","MT3","DAB1","C12orf57","CASP3","EFNB2","CREB1","STMN1","RAB3A","TRIM32","BSG","ROBO1","TUBB3","SIPA1L1","SLC9A6","FKBP1B","SEMA3A","LMTK2","SLITRK1","SPAST","ITGB1","SLITRK5","KIAA0319","TIAM2","GSK3B","KIF5B","NRXN1","EPHA7","PLXNC1","CDK5R2","ISLR2","INPP5F","EFNA5","PLPPR4","AMIGO1","SEMA6B","PTCH1","RAB10","HSP90AA1","DPYSL2","LGI1","APLP1","DCC","KIF5C","RTN4","EDNRA","THY1","ALCAM","PTN","TUBB2B","MAP2","NREP","TMEFF1","CDH2","ZDHHC17","FZD3","SLC25A46","TNFRSF21","B4GALT6","SLITRK4","GDI1","PTEN","CNTN4","NDN","RAC1","HSP90AB1","PALLD","RNF6","CDK5R1","RPL24","PTPRZ1","RELN","OLFM1","CCK","FLRT2","CNR1","EDNRA","ADARB1","ALS2","SPAST","NTNG1","ADNP","APLP1","AMIGO1","UCHL1","SLITRK3","NTRK2","PTPRZ1","RGMA","LMTK2","NCAM2","MAP1B","TOP2B","SPART","SEMA3C","TAOK2","PICALM","NFIB","MAP2K2","B4GAT1","CNTN1","DLX5","FZD3","APP","TUBB3","OLFM1","MAP1A","APLP2","MAP2","KIF5A","PTEN","CDH2","GDI1","BSG","TMEFF1","PTN","LGI1","ADCY1","ATL1","FEZ2","PLXNC1","SEMA6A","SVBP","TNFRSF21","EPHA5","FBXO45","PLPPR4","EPHA7","DCLK1","RAB21","NPTN","ZDHHC17","NREP","KIF5B","RAB3A","FLRT2","JAK2","RTN4","SLC25A46","WNT5A","ISLR2","RAC1","SLITRK5","TUBB2B","CSPG5","RNF6","ETV1","THY1","B4GALT5","B4GALT6","STMN1","C12orf57","CDK5R1","NOTCH2","ACTB","SLC9A6","FOXG1","HSP90AA1","NDN","INPP5F","HSP90AB1","CNR1","SLITRK4","EFNB2","ITGB1","CCK","RPL24","FGF13","ROBO1","SEMA6C","EPHA6","VASH2","TOP2B","RNF6","FEZ2","SEMA4F","PTPN11","NOVA2","MAP1A","CREB1","SLC9A6","NCAM2","TUBB2B","TNFRSF21","EPHA4","EPHA7","CDK5","MCF2","IFRD1","NR2E1","GSK3B","CHRNB2","ROBO2","RAB3A","ADCY1","SLC25A46","SLITRK1","BRAF","ADNP","NDN","RTN4","ANOS1","UNC5C","PTN","DAB1","RAB21","ITGB1","COBL","SMURF1","ALCAM","ULK2","KIF5B","NPTN","WNT5A","SLITRK4","C9orf72","EPHA3","EFNB2","ZDHHC17","NGF","PALLD","EPHA5","LRRC4C","PTPRZ1","APP","ACTB","MYCBP2","APLP2","C12orf57","ANK3","RAC1","CNTN1","CDK5R1","INPP5F","SLITRK5","FZD3","FOXG1","SEMA3A","OLFM1","ETV1","STMN1","THY1","HSP90AB1","RELN","PICALM","RPL24","HSP90AA1","BCL2","CDH11","CCK","FLRT2","CNR1","GSK3B","CTNNA2","CASP3","SEMA6D","EPHA4","SPAST","CHRNB2","CNP","STXBP1","YTHDF1","EPHA5","RTN4","XK","TUBB3","PLXNA2","NPTN","TUBB2B","MAP1A","CNTN4","CSPG5","B4GALT5","RAB21","B4GAT1","ZDHHC17","KIF5C","JAK2","CREB1","TRPC5","NTRK2","EPHA3","APP","MAP2K2","ACTB","PICALM","APLP1","SLITRK4","EFNB2","CNTN6","KIAA0319","SLC25A46","STK11","INPP5F","GDI1","APLP2","FZD3","NOTCH2","FGF13","BCL11B","OLFM1","BCL2","SLITRK1","FBXO45","FLRT2","UCHL1","UNC5C","DCC","KIF5B","ADCY1","DHFR","CDK5R1","EPHB6","NFIB","AMIGO1","DISC1","CDH11","RAC1","RELN","ITGB1","S100A6","RAB3A","C12orf57","PLPPR4","RNF6","SLITRK5","SVBP","CCK","SEMA6A","TMEFF1","STMN1","ETV1","NDN","HSP90AB1","RPL24","HSP90AA1","NDEL1","PTN","MBP","CDH11","TCTN1","SEMA6B","TUBB3","SLITRK2","RNF6","LAMA5","CYFIP1","GSK3B","MYCBP2","SOS1","CNTN6","RYK","SRF","PPP3CB","CNR1","SLITRK4","TBC1D24","STMN1","TOP2B","UCHL1","EFNA5","IGF1R","EPHA3","ADCY1","SLIT2","NPTN","KIAA0319","HSP90AA1","LGI1","FOXG1","C9orf72","DRAXIN","DPYSL5","LRTM2","NOVA2","NREP","RAB21","MEGF8","BRAF","PTPRZ1","PAX6","TRIO","CTNNA2","EXT1","CERS2","THY1","GDI1","SLITRK3","GAP43","CDH2","SIPA1L1","APP","PSEN1","SIAH1","CREB1","ARX","BCL2","NRCAM","SHTN1","NRXN3","SPART","SLITRK5","SLITRK1","CDKL3","TIAM2","NTRK2","LRRC4C","UNC5D","IFRD1","CHN1","FGF13","LRTM1","B4GAT1","PAK1","DBN1","SEMA6D","EPHB1","NRP1","DPYSL2","RELN","FZD3","FEZ2","HSP90AB1","KLF7","INPP5F","PICALM","EFNB2","FLRT2","NCAM2","CCK","KIF5C","EPHA4","RPL24","KIF5B","SEMA6D","CAMSAP2","MAP2","SPRY3","MAP1S","LAMA3")
```

### Extended data fig. 17
```{r fig.height=15, fig.width=15}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(viridis)
library(ggrepel)
library(gridExtra)
library(cowplot) 

# Specify genes of interest
genes_of_interest <- axon_development_genes

#Get higher level annotation de genes

cao$estimateDEPerCellType(cell.groups = annotation_n$l3,name = "de_l3")
de_l3 <- cao$test.results$de_l3
```


```{r fig.height=15, fig.width=15}
# Create a combined data frame for plotting
plot_data <- data.frame()

for (celltype in names(de_l3)) {
  if (celltype != "TBD") { # Exclude cell type "TBD"
    res <- as.data.frame(de_l3[[celltype]]$res) # Ensure it's a plain dataframe
    res$celltype <- celltype
    plot_data <- rbind(plot_data, res)
  }
}

# Ensure 'Gene' column is used for gene names
plot_data$gene <- plot_data$Gene

# Convert necessary columns to appropriate types
plot_data$log2FoldChange <- as.numeric(plot_data$log2FoldChange)
plot_data$padj <- as.numeric(plot_data$padj)

# Filter for genes of interest
plot_data <- plot_data %>%
  filter(gene %in% genes_of_interest)

# Calculate absolute log2FoldChange for color scaling
plot_data <- plot_data %>%
  mutate(abs_log2FoldChange = abs(log2FoldChange))

# Function to create volcano plot for each cell type
create_volcano_plot <- function(celltype_data, celltype_name, color_limits) {
  # Select top 20 most changed genes based on both padj and log2FoldChange, ensuring log2FoldChange >= 1
  top_genes <- celltype_data %>%
    filter(abs(log2FoldChange) >= 1) %>%
    arrange(padj, desc(abs_log2FoldChange)) %>%
    slice_head(n = 20)
  
  # Include all outliers
  outliers <- celltype_data %>%
    filter(log2FoldChange > 4 | log2FoldChange < -4)
  
  # Combine top genes and outliers, ensuring no duplicates
  labeled_genes <- bind_rows(top_genes, outliers) %>%
    distinct(gene, .keep_all = TRUE)
  
  ggplot(celltype_data, aes(x = log2FoldChange, y = -log10(padj), color = abs_log2FoldChange)) +
    geom_point(alpha = 0.6) +
    geom_text_repel(data = labeled_genes, 
                    aes(label = gene), 
                    size = 3, fontface = "bold", box.padding = 0.5, point.padding = 0.5, segment.color = "grey50", max.overlaps = 50) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red", linewidth = 0.5) +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", linewidth = 0.5) +
    scale_color_viridis(option = "H", limits = color_limits) +
    labs(title = celltype_name,
         x = "log2 Fold Change",
         y = "-log10(padj)",
         color = "Absolute log2 Fold Change") +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5),
      legend.position = "right"
    )
}

# Calculate global color limits
color_limits <- c(1, max(plot_data$abs_log2FoldChange, na.rm = TRUE))

# Create a list to store plots
volcano_plots <- list()

# Generate volcano plots for each cell type
for (celltype in unique(plot_data$celltype)) {
  celltype_data <- plot_data %>% filter(celltype == !!celltype)
  volcano_plots[[celltype]] <- create_volcano_plot(celltype_data, celltype, color_limits)
}

# Extract the legend from one of the plots
legend <- get_legend(volcano_plots[[1]])

# Remove legends from individual plots
volcano_plots <- lapply(volcano_plots, function(plot) plot + theme(legend.position = "none"))

# Arrange all volcano plots in one panel with shared legend
combined_plot <- plot_grid(plotlist = volcano_plots, ncol = 3)
final_plot <- plot_grid(combined_plot, legend, ncol = 2, rel_widths = c(4, 0.5))

# Display the final plot
print(final_plot)
p <- final_plot
ggsave(plot = p, filename = "figures/cacoa/volcano_plot_axon_develop_GSEA.pdf",dpi = 300,width = 16,height = 14)
```


# Gene Set analyses
## Genes are fetched from GSEA of grepl "electron" and uses genes from electron transport chain complexes and other genes in the related pathways.
## Load Seurat object for neurons
```{r}
seurat_obj <- readRDS("../../TSC2024_data_files/5.neurons/seurat_neurons.rds")
```


## Defined genesets for electron transport chain complexes and associated genes
```{r}
# Load necessary libraries
library(Seurat)
library(ggplot2)
library(dplyr)
library(ggpubr)
library(cowplot)

# Define your genesets once
genesets <- list(
  complex_I_genes = c("NDUFA1", "NDUFA2", "NDUFA3", "NDUFA4", "NDUFA5", "NDUFA6", "NDUFA7", "NDUFA8", "NDUFA9", "NDUFA10", "NDUFA11", "NDUFA12", "NDUFB1", "NDUFB2", "NDUFB3", "NDUFB4", "NDUFB5", "NDUFB6", "NDUFB7", "NDUFB8", "NDUFB9", "NDUFB10", "NDUFB11", "NDUFC1", "NDUFC2", "NDUFS1", "NDUFS2", "NDUFS3", "NDUFS4", "NDUFS5", "NDUFS6", "NDUFS7", "NDUFS8", "NDUFV1", "NDUFV2", "NDUFV3", "MT-ND1", "MT-ND2", "MT-ND3", "MT-ND4", "MT-ND4L", "MT-ND5", "MT-ND6")  ,
  complex_II_genes = c("SDHA", "SDHB", "SDHC", "SDHD"),
  complex_III_genes = c("UQCR11", "UQCRB", "UQCRC1", "UQCRC2", "UQCRFS1", "UQCRH", "UQCRQ", "CYC1", "MT-CYB"),
  complex_IV_genes = c("COX4I1", "COX4I2", "COX5A", "COX5B", "COX6A1", "COX6A2", "COX6B1", "COX6B2", "COX6C", "COX7A1", "COX7A2", "COX7A2L", "COX7B", "COX7B2", "COX7C", "COX8A", "COX8C", "MT-CO1", "MT-CO2", "MT-CO3"),
  complex_V_genes = c("ATP5F1A", "ATP5F1B", "ATP5F1C", "ATP5F1D", "ATP5F1E", "ATP5MC1", "ATP5MC2", "ATP5MC3", "ATP5ME", "ATP5MF", "ATP5MG", "ATP5PD", "ATP5PO", "ATP5PB", "ATP5PF", "ATP5S", "ATP5SL", "MT-ATP6", "MT-ATP8", "ATP5IF1", "ATP5A1-AS1"),
  other_genes = c("SNCA", "ISCU", "CYCS", "AFG1L", "PARK7", "DNAJC15", "PINK1", "COQ9", "SLC25A51", "DLD", "DGUOK", "CHCHD2", "GHITM", "COA6", "BID", "CCNB1", "AKR1B1", "DEGS1", "GLRX2", "ADH5", "CYB5A", "SRD5A1", "FDX1", "SURF1", "FDX2", "CIAPIN1", "ASPH", "ETFB", "P4HA2", "QDPR", "GSR", "ME2", "POR", "DHRS3", "ETFA", "NOX4", "ETFDH", "PPARGC1A", "PUM2", "GPD2", "IMMP2L", "ETFRF1", "SLC25A22", "SOD2", "WDR93", "PDIA5", "PLEC", "SLC25A13", "POLG2", "CYB561", "ENOX2", "SLC25A12")
)
```

## Individual genesets for celltypes
### Extended data fig. 12

```{r}
# Filter out the specific cell types "TBD" and "Neu_dev_THY1"
seurat_obj_filtered <- subset(seurat_obj, annol3 != "TBD" & annol3 != "Neu_dev_THY1")

# Function to create violin plots for a geneset, summarized by average expression per cell
plot_geneset_violin <- function(seurat_obj, geneset, geneset_name) {
  geneset <- intersect(geneset, rownames(seurat_obj))  # Ensure genes are in the Seurat object
  
  if(length(geneset) > 0) {
    # Calculate the average expression for each geneset per cell
    avg_expr <- colMeans(GetAssayData(seurat_obj, assay = "RNA", layer = "data")[geneset, , drop = FALSE])
    
    # Convert avg_expr to a dataframe
    avg_expr_df <- data.frame(avg_expr)
    
    # Add the average expression as a metadata column
    seurat_obj <- AddMetaData(seurat_obj, metadata = avg_expr_df, col.name = geneset_name)
    
    # Ensure the 'condition' column exists
    if(!"condition" %in% colnames(seurat_obj@meta.data)) {
      stop("The 'condition' column is not found in the Seurat object metadata.")
    }
    # Filter out the specific cell types "TBD" and "Neu_dev_THY1"
    seurat_obj_filtered <- subset(seurat_obj, annol3 != "TBD" & annol3 != "Neu_dev_THY1")
    
    # Create a data frame for ggplot
    plot_data <- FetchData(seurat_obj_filtered, vars = c(geneset_name, "annol3", "condition"))
    
    # Set the factor levels for annol3 to ensure the desired order
    plot_data$annol3 <- factor(plot_data$annol3, levels = c("L2_CUX2", "L2_3_CUX2", "L4_RORB", "L5_6_FEZF2_LRRK1", "L5_6_FEZF2_TLE4", "L5_6_THEMIS","ID2","VIP","PVALB","SST"))
    
    # Perform statistical test
    stat_test <- compare_means(formula = as.formula(paste(geneset_name, "~ condition")), data = plot_data, group.by = "annol3", method = "wilcox.test")
    
    # Adjust p-values using Bonferroni method
    stat_test$p.adj <- p.adjust(stat_test$p, method = "bonferroni")
    
    # Save the expression data along with the gene set
    expression_data <- GetAssayData(seurat_obj_filtered, assay = "RNA", layer = "data")[geneset, , drop = FALSE]
    #write.csv(expression_data, file = paste0("figures/cacoa/gene_set_tests/", geneset_name, "_expression_data.csv"), row.names = TRUE)
    
    # Create the violin plot
    p <- ggplot(plot_data, aes(x = annol3, y = !!sym(geneset_name), fill = condition)) +
      geom_violin(trim = FALSE) +
      geom_boxplot(width = 0.1, position = position_dodge(0.9), outlier.size = 0.5) +
      ggtitle(paste(geneset_name, "Expression")) + 
      theme(plot.title = element_text(hjust = 0.5)) +
      scale_fill_manual(values = alpha(c("Control" = "dodgerblue2", "TSC2" = "deeppink3"), .5)) +
      xlab("Cell Type") +  # Set x-axis title
      ylab("Normalized expression") + # Set y-axis title
      theme_minimal() +
      stat_compare_means(aes(group = condition), method = "wilcox.test", label = "p.signif")  # Add statistical test with asterisks
    p <- p %>% rasterize(dpi=600)
    # Save the plot
    ggsave(filename = paste0("figures/cacoa/gene_set_tests/", geneset_name, "_violin_plot.pdf"), width = 14, height = 6, dpi = 600, plot = p)
    
    # Save the table with p-values and adjusted p-values
    #write.csv(stat_test, file = paste0("figures/cacoa/gene_set_tests/", geneset_name, "_stat_test.csv"), row.names = FALSE)
    
    # Return the plot and table
    return(list(plot = p, stat_test = stat_test))
  } else {
    message(paste("No genes from the geneset", geneset_name, "found in the Seurat object"))
    return(NULL)
  }
}

# Plot each geneset and collect the plots
plots <- list()
  for (geneset_name in names(genesets)) {
  result <- plot_geneset_violin(seurat_obj_filtered, genesets[[geneset_name]], geneset_name)
  if (!is.null(result)) {
    plots[[geneset_name]] <- result$plot
  }
}

# Combine all plots into one
combined_plot <- plot_grid(plotlist = plots, ncol = 1)

# Save the combined plot
ggsave(filename = "figures/cacoa/gene_set_tests/combined_violin_plot.pdf", plot = combined_plot, width = 14, height = 24, dpi = 200)

# Print the combined plot
print(combined_plot)

```

## All genesets in one graph:
### Fig. 3d
```{r}
# Function to create violin plots for a geneset, summarized by average expression per cell
plot_geneset_violin <- function(seurat_obj, geneset, geneset_name) {
  geneset <- intersect(geneset, rownames(seurat_obj))  # Ensure genes are in the Seurat object
  
  if(length(geneset) > 0) {
    # Calculate the average expression for each geneset per cell
    avg_expr <- colMeans(GetAssayData(seurat_obj, assay = "RNA", layer = "data")[geneset, , drop = FALSE])
    
    # Convert avg_expr to a dataframe
    avg_expr_df <- data.frame(avg_expr)
    
    # Add the average expression as a metadata column
    seurat_obj <- AddMetaData(seurat_obj, metadata = avg_expr_df, col.name = geneset_name)
    
    # Ensure the 'condition' column exists
    if(!"condition" %in% colnames(seurat_obj@meta.data)) {
      stop("The 'condition' column is not found in the Seurat object metadata.")
    }
    
    # Filter out the specific cell types "TBD" and "Neu_dev_THY1"
    seurat_obj_filtered <- seurat_obj
    
    # Create a data frame for ggplot
    plot_data <- FetchData(seurat_obj_filtered, vars = c(geneset_name, "condition"))
    plot_data$geneset <- geneset_name  # Add a column for the geneset name
    plot_data$avg_expr <- plot_data[[geneset_name]]  # Ensure avg_expr column is present
    
    return(plot_data)
  } else {
    message(paste("No genes from the geneset", geneset_name, "found in the Seurat object"))
    return(NULL)
  }
}

# Combine data for all genesets
all_plot_data <- bind_rows(lapply(names(genesets), function(geneset_name) {
  plot_geneset_violin(seurat_obj, genesets[[geneset_name]], geneset_name)
}))

# Create the combined violin plot
combined_plot <- ggplot(all_plot_data, aes(x = condition, y = avg_expr, fill = condition)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1, position = position_dodge(0.9), outlier.size = 0.5) +
  facet_wrap(~ geneset, scales = "fixed", nrow = 1) +  # Use fixed y-axis and one row
  ggtitle("Expression of Electron transport chain genesets") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values = alpha(c("Control" = "dodgerblue2", "TSC2" = "deeppink3"), .5)) +
  xlab("Condition") +  # Set x-axis title
  ylab("Normalized expression") + # Set y-axis title
  theme_minimal() +
  stat_compare_means(aes(group = condition), method = "wilcox.test", label = "p.signif",label.x.npc = "center", label.y.npc = "top")  # Add statistical test with asterisks


combined_plot <- combined_plot %>% rasterize(dpi=600)
# Save the combined plot
ggsave(filename = "figures/cacoa/gene_set_tests/all_genesets_combined_violin_plot.pdf", width = 14, height = 6, dpi = 400, plot = combined_plot)

# Display the combined plot
print(combined_plot)
```

#VIP Violin plots CXCL14 in all VIP cell types:
### Extended data fig. 14 
```{r}
# Load necessary libraries
library(Seurat)
library(ggplot2)
library(dplyr)
library(ggpubr)

# Define the cell types of interest
cell_types <- c("VIP_ABI3BP", "VIP_CBLN1", "VIP_CRH", "VIP_NRG1", "VIP_SEMA3C", "VIP_TRPM3_SSTR1", "VIP_TYR")

# Ensure the 'condition' column exists
if(!"condition" %in% colnames(seurat_obj@meta.data)) {
  stop("The 'condition' column is not found in the Seurat object metadata.")
}

# Filter the Seurat object to include only the specified cell types
seurat_obj_filtered <- subset(seurat_obj, annol4 %in% cell_types)

# Extract the expression data for the gene "CXCL14"
cxcl14_data <- FetchData(seurat_obj_filtered, vars = c("CXCL14", "annol4", "condition"))

# Create the violin plot
violin_plot <- ggplot(cxcl14_data, aes(x = condition, y = CXCL14, fill = condition)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1, position = position_dodge(0.9), outlier.size = 0.5) +
  facet_wrap(~ annol4, scales = "fixed", nrow = 1) +  # Use fixed y-axis and one row
  ggtitle("CXCL14 Expression in VIP Cell Types") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values = alpha(c("Control" = "dodgerblue2", "TSC2" = "deeppink3"), .5)) +
  xlab("Condition") +  # Set x-axis title
  ylab("Normalized expression") + # Set y-axis title
  theme_minimal() +
  stat_compare_means(aes(group = condition), method = "wilcox.test", label = "p.signif",label.x.npc = "center", label.y.npc = "top")  # Add statistical test with asterisks

# Save the violin plot
ggsave(filename = "figures/cacoa/cxcl14_violin_plot.pdf", width = 14, height = 6, dpi = 400, plot = violin_plot)

# Display the violin plot
print(violin_plot)

```
#VIP Violin plots CXCL14 in VIP cells in general:
```{r}
# Load necessary libraries
library(Seurat)
library(ggplot2)
library(dplyr)
library(ggpubr)

# Define the cell types of interest
cell_types <- c("VIP_ABI3BP", "VIP_CBLN1", "VIP_CRH", "VIP_NRG1", "VIP_SEMA3C", "VIP_TRPM3_SSTR1", "VIP_TYR")

# Ensure the 'condition' column exists
if(!"condition" %in% colnames(seurat_obj@meta.data)) {
  stop("The 'condition' column is not found in the Seurat object metadata.")
}

# Filter the Seurat object to include only the specified cell types
seurat_obj_filtered <- subset(seurat_obj, annol4 %in% cell_types)

# Extract the expression data for the gene "CXCL14"
cxcl14_data <- FetchData(seurat_obj_filtered, vars = c("CXCL14", "annol4", "condition"))

# Create the violin plot
violin_plot <- ggplot(cxcl14_data, aes(x = condition, y = CXCL14, fill = condition)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1, position = position_dodge(0.9), outlier.size = 0.5) +
  ggtitle("CXCL14 Expression in VIP Cell Types") + 
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values = alpha(c("Control" = "dodgerblue2", "TSC2" = "deeppink3"), .5)) +
  xlab("Condition") +  # Set x-axis title
  ylab("Normalized expression") + # Set y-axis title
  theme_minimal() +
  stat_compare_means(aes(group = condition), method = "wilcox.test", label = "p.signif",label.x.npc = "center", label.y.npc = "top")  # Add statistical test with asterisks

# Save the violin plot
ggsave(filename = "figures/cacoa/cxcl14_violin_plot_collapsed.pdf", width = 6, height = 6, dpi = 400, plot = violin_plot)

# Display the violin plot
print(violin_plot)

```



# Celltype proportion plot, conditions
```{r}
identifier <- read.csv2("../identifier_list.csv")
identifier$X <- NULL
cms_neurons <- readRDS("cms_neurons.rds")
names(cms_neurons) <- identifier$pseudo

```
```{r}
# Initialize a list to store the counts for each matrix
individual_counts <- list()

# Loop through each matrix in cms
for (i in seq_along(cms_neurons)) {
  # Get the cell names for the current matrix
  cell_names <- colnames(cms_neurons[[i]])
  
  # Ensure the cell names are in annotation_n$l4
  matching_cells <- cell_names[cell_names %in% names(annotation_n$l4)]
  
  # Count the number of cells based on the levels in annotation_n$l4
  cell_counts <- table(annotation_n$l4[matching_cells])
  
  # Store the counts in the list
  individual_counts[[names(cms_neurons)[i]]] <- cell_counts
}

# Convert the list to a data frame
df_counts <- do.call(rbind, lapply(individual_counts, as.data.frame))

# Add a column for the sample names
df_counts$Sample <- rep(names(individual_counts), sapply(individual_counts, nrow))

# Calculate the relative amounts
df_counts$RelativeAmount <- df_counts$Freq / ave(df_counts$Freq, df_counts$Sample, FUN = sum)

# Reorder columns to have Sample first
df_counts <- df_counts[, c("Sample", "Var1", "Freq", "RelativeAmount")]

# Rename columns for clarity
colnames(df_counts) <- c("Sample", "CellType", "Freq", "RelativeAmount")

# Print the combined data frame
print(df_counts)
write.csv2(df_counts, "Cell_type_proportions_neurons.csv")
```

## Collapse samples
```{r}
# Add a new column to group by prefix
df_counts$Prefix <- ifelse(grepl("^control", df_counts$Sample), "control",
                           ifelse(grepl("^tsc", df_counts$Sample), "tsc", df_counts$Sample))

# Summarize the data by collapsing by Prefix and CellType
df_collapsed <- df_counts %>%
  group_by(Prefix, CellType) %>%
  summarise(Freq = sum(Freq), .groups = 'drop')

# Calculate the relative frequency per cell type
df_collapsed <- df_collapsed %>%
  group_by(CellType) %>%
  mutate(RelativeFrequency = Freq / sum(Freq)) %>%
  ungroup()

# Rename columns for clarity
colnames(df_collapsed) <- c("Condition", "CellType", "Freq", "RelativeFrequency")

# Print the collapsed data frame
print(df_collapsed)
```

```{r}

# Create the bar plot
ggplot(df_collapsed, aes(x = CellType, y = RelativeFrequency, fill = Condition)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Relative Cell Type Proportions",
       x = "Sample",
       y = "Proportion (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = -45,hjust = 0))

filename = "figures/relative_cell_type_proportions_neurons_condition.pdf"
pdf.width = 10
pdf.height = 6
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```
## Normalizing for cell numbers per condition 
```{r}
# Define the number of samples for each condition
control_samples <- 6
tsc_samples <- 11

# Normalize the frequencies by the number of samples
df_collapsed <- df_collapsed %>%
  mutate(NormalizedFreq = ifelse(Condition == "control", Freq / control_samples, 
                                 ifelse(Condition == "tsc", Freq / tsc_samples, Freq)))

# Recalculate the relative frequency based on the normalized frequencies
df_collapsed <- df_collapsed %>%
  group_by(CellType) %>%
  mutate(NormalizedRelativeFrequency = NormalizedFreq / sum(NormalizedFreq)) %>%
  ungroup()

# Print the final data frame
print(df_collapsed)

# Optionally, save the normalized data frame to a CSV file
write.csv2(df_collapsed, "Cell_type_proportions_neurons_glia_collapsed_normalized.csv", sep = ",", dec=".")

```

## Collapsed samples with normalization:
### Fig. 1i
```{r}
# Create the bar plot
ggplot(df_collapsed, aes(x = CellType, y = NormalizedRelativeFrequency, fill = Condition)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Normalized Relative Cell Type Proportions",
       x = "Sample",
       y = "Proportion (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = -45,hjust = 0))

filename = "figures/relative_cell_type_proportions_neurons_condition_normalized_to_sample_number.pdf"
pdf.width = 10
pdf.height = 6
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```

## Neu_dev_THY1 dotplot
### Extended data fig. 8a

```{r fig.height=2, fig.width=15}
# Subset the Seurat object to include only the "Neu_dev_THY1" cell type
Idents(seurat_obj) <- "annol4"
seurat_obj_subset <- subset(seurat_obj, idents = "Neu_dev_THY1")

# Define the markers
markers <- c("RBFOX3", "SLC17A7", "GAD1", "CUX2", "RORB", "FEZF2", "THEMIS", "ID2", "VIP", "PVALB", "SST","CALM2","NRGN","CHN1","SNAP25","THY1")

# Create the DotPlot for the subsetted Seurat object
DotPlot(seurat_obj_subset, features = markers, cols = c("blue", "red"), dot.scale = 8) +
  RotatedAxis() +
  xlab(element_blank()) +
  ylab(element_blank())

filename = "figures/cacoa/Neu_dev_THY1_identity_markers.pdf"
pdf.width = 10
pdf.height = 4
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```
### Extended data fig. 8b

```{r}
seurat_obj_subset <- subset(seurat_obj, idents = "Neu_dev_THY1")
# Developmental (age-associated) neuronal markers (2-month-old PFC)
dev_markers <- c(
  "RBFOX3",
  "NEUROD6",
  "SATB2",
  "SYT1",
  "MAP2",
  "STMN2",
  "DCX",
  "BCL11B",
  "THY1",
  "TUBB3",
  "SYN1"
)

# TSC pathology-associated cortical tuber markers
tsc_pathology_markers <- c(
  "VIM",
  "NES",
  "GFAP",
  "CD44",
  "NEFH",
  "RPS6",
  "PROM1",
  "CD34",
  "PDGFRA"
)

# Create the DotPlot for the subsetted Seurat object
DotPlot(seurat_obj_subset, features = c(dev_markers,tsc_pathology_markers), cols = c("blue", "red"), dot.scale = 8) +
  RotatedAxis() +
  xlab(element_blank()) +
  ylab(element_blank())

filename = "figures/cacoa/Neu_dev_THY1_early+TSC_markers.pdf"
pdf.width = 10
pdf.height = 4
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```


# CUX2,RORB, FEZF2 excitatory neuron dotplot for RNAscope
### Fig. 7b
```{r fig.height=4, fig.width=7}

excitatory_neurons <- c(
  "L2_CUX2_LAMP5_PDGFD", "L3_CUX2_SPOCK3_PRSS12","L2_3_CUX2_FREM3_NTNG1",
  "L4_RORB_SCHLAP1_ARHGAP15", "L4_RORB_SCHLAP1_MET", "L4_RORB_SCHLAP1_MME",
  "L5_6_FEZF2_LRRK1_PCP4", "L5_6_FEZF2_LRRK1_SEMA3E", "L5_6_FEZF2_TLE4_ABO","L5_6_FEZF2_TLE4_HTR2C", "L5_6_FEZF2_TLE4_SCUBE1",
  "L5_6_THEMIS_NR4A2_NTNG2", "L5_6_THEMIS_SEMA3A"
)


#Idents(seurat_obj) <- seurat_obj$annol4
seurat_obj_subset <- subset(seurat_obj, idents = excitatory_neurons)

# Define the markers
markers <- c("CUX2", "RORB", "FEZF2")

# Create the DotPlot for the subsetted Seurat object
DotPlot(seurat_obj_subset, features = markers, cols = c("blue", "red"), dot.scale = 8) +
  coord_flip() +
  RotatedAxis() +
  xlab(element_blank()) +
  ylab(element_blank())

filename = "figures/cacoa/Excitatory_RNAscope_markers.pdf"
pdf.width = 7
pdf.height = 4
dev.copy2pdf(file=filename, width = pdf.width, height = pdf.height)
```
```{r}
sessionInfo()
```


